### 2015-09-23 draft1leslieandkenny.Rnw
Getting started

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}

\setlength{\textwidth}{6.5in}
\setlength{\textheight}{9in}
\setlength{\oddsidemargin}{0in}
\setlength{\evensidemargin}{0in}
\setlength{\topmargin}{-1.5cm}
\setlength{\maxwidth}{6.5in}

\title{\vspace{-2.0cm}The effect of nitrogen application time on infection and development of the Wheat Streak Mosaic Virus (WSMV)}
\author{Leslie Gains-Germain, Kenny Flagg}
\date{Spring 2015}

\begin{document}
<<setup, include=FALSE, cache=FALSE,echo=FALSE>>=
require(knitr)
opts_chunk$set(fig.width = 10, fig.height = 4, out.width = '\\linewidth',
               out.height = '0.4\\linewidth', dev = 'pdf',
               concordance = TRUE,
               tidy.opts = list(width.cutoff = 60, keep.blank.line=FALSE),
               size = 'footnotesize')
options(replace.assign = TRUE, width = 112, digits = 3, max.print = '72',
        show.signif.stars = FALSE)

require(xtable)
#require(foreign)
#require(arm)
require(tabplot)
require(dplyr)
require(car)
require(lme4)
require(ggplot2)
require(lattice)
require(effects)
@

<<data,echo=FALSE>>=
# Read and set up the data
fert2 <- read.csv('fert2.csv')

# Correctly order nitrogen application levels
fert2$n.trt <- relevel(fert2$n.application, 'FALL')

# Create a row factor
fert2$row <- with(fert2, interaction(variety, block))

# Combine nitrogen and inoc/cntl
fert2$n.status <- with(fert2, interaction(n.trt, status))
@

\maketitle
\tableofcontents

\newpage

\section{Introduction}
\noindent

\section{Statistical Experience and Assistance Needed}
\noindent Nar B. Ranabhat is a PhD candidate in the Department of Land Resources and Environmental Science. He is currently working on several projects regarding the spread and development of the Wheat Streak Mosaic Virus (WSMV) in Winter Wheat. Nar has taken Statistics $511$ and $512$ here at Montana State University, and he is currently enrolled in Mixed Models. Nar has asked for our help in building a model, fitting it in R, and interpreting the results.

\section{Objectives and Questions}
\noindent Nar is interested in the effect of variety, nitrogen application time, and inoculation on the probability of virus infection in Winter Wheat. He is primarily interested in the effect of nitrogen application time, and he wants to assess evidence for all possible two and three way interactions. Nar's goal is to select the simplest possible model, and then use this model to describe the effects of the above variables. \\

We address the following questions in this report:
\begin{itemize}
\item Is there evidence that the interaction between nitrogen application and variety differs between inoculated and control plots?
\item Is there evidence that the nitrogen application effect varies across varieties? Is there evidence that the nitrogen application effect differs between inoculated and control?
\item Is there evidence that the variety effect differs between inoculated and control?
\item Conditional on the results to the above questions, how do we appropriately describe the estimated effects for variety, nitrogen application time, and inoculation on the probability of virus infection?
\end{itemize}

\section{Study Design/Data collection}

Nar conducted this experiment in a MSU field at the base of the Bridger Mountains. He chose one area in the field and divided it into six $31.5$ by $27.5$ meter blocks. The blocks were then divided into five rows, with $5$ meters of space between each row. Each row was randomly assigned to a variety of Winter Wheat with separate randomizations in each block. The rows then were divided into six $1.5$ by $5$ meter plots. Each plot was randomly assigned to one of six combinations of nitrogen application time and inoculation status, with separate randomizations in each row. The combinations were fall/inoculated, fall/control, early spring/inoculated, early spring/control, late spring/inoculated, and late spring/control. All plots were planted in the {\bf FALL?} and allowed to grow over the winter. The following {\bf September?}, $30$ plants were chosen from each plot to be taken to the lab and tested for presence of the virus. The selection of the plants from each plot was not random nor technically systematic. The {\bf undergraduate?} technicians were instructed to collect a sample of plants spread evenly throughout the field. {\bf different undergraduates doing picking? picking done all at the same time? is there visual evidence of the disease?} A total of $180$ plants were collected. The plants were then sent to the lab and screened for the virus via the ELISA procedure. This is a three year study. The data are collected for the first year, and the second year plants are waiting to be analyzed in the lab {\bf describe how they are stored during waiting time}. Planting for the third year will begin this fall {\bf I think?}.  \\

\noindent Plots that were assigned to the inoculated treatment had five infected plants transplanted to the middle of the plot. These plants were infected {\bf as adults?} in the greenhouse by clipping an infected leaf to the healthy plant. The Wheat Streak Mosaic Virus is transmitted by the wheat curl mite, tiny organisms that are nearly invisible to the naked eye. They can easily move from leaf to leaf on a plant, and they are transported from plant to plant by the wind  (Sloderbeck $2008$). It is our understanding that once the infected mites were introduced to the study area, they were considered to be everywhere in the air and that all plants in the field were exposed to the infected mites.

\section{Recommendations}

\subsection{Binomial Generalized Linear Mixed Model}

\subsection{Graphical Data Exploration}

<<interactionplot,echo=FALSE>>=
qplot(x = n.trt, y = infected, geom = 'boxplot', color = status,
      facets = .~variety, data = fert2) +
  theme(axis.text.x = element_text(angle = 90))
@

<<heatmap,echo=FALSE>>=
## Heatmap with meaningful ordering

# Order the dataset
#fert.ordered <- arrange(fert2, block, variety, status, n.trt)
fert.ordered <- arrange(fert2, variety, block, status, n.trt)

# Create a matrix of the arranged responses
infected.arranged <- matrix(fert.ordered$infected, ncol = 6, byrow = TRUE)

# Set up two panels, right one for a legend
layout(t(1:2), widths = c(9, 1))

# Plot the heatmap, with zeros in black and segments separating the blocks
par(mar = c(5, 10, 6, 2))
image(x = seq(0.5, 5.5, 1/6), y = 1:6, z = infected.arranged,
      col = c('black', heat.colors(30)), xlab = '', ylab = '',
      zlim = c(0, 30), yaxt = 'n', xaxt = 'n')
title('Infection Counts', line = 4)
segments(x0 = 1.5:4.5, y0 = 0.5, y1 = 6.5, col = 'white')
axis(3, labels = rep(levels(fert2$block), 5),
     at = seq(7/12, 5 + 5/12, 1/6), cex.axis = 0.75)
axis(2, labels = levels(fert2$n.status), at = 1:6, las = 2)
axis(1, labels = levels(fert2$variety), at = 1:5)

# Legend
par(mar = c(5, 1, 6, 2))
image(y = -0.5:30.5, z = matrix(0:30, nrow = 1), axes = FALSE, ylab = '',
      col = c('#000000', heat.colors(30)), zlim = c(0, 30))
title('Legend', line = 1.5)
axis(4)
@

<<tabplot,echo=FALSE,warning=FALSE,message=FALSE>>=
tableplot(fert.ordered, select = c(infected, variety, block, n.trt, status),
          nBins = 30, sortCol = variety)
@

The \verb|itableplot| function in the \verb|tabplot| package is another
useful tool for visualizing the data and identifying patterns.

\subsection{Convergeance Warning Messages}

<<models,cache=TRUE>>=
glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status +
                    (1|block/row),
                  control = glmerControl(optimizer = 'bobyqa'),
                  family = binomial, data = fert2)
summary(glmm3way)

glmm2way <- glmer(cbind(infected, total) ~ variety*n.trt + variety*status +
                    n.trt*status + (1|block/row),
                    control = glmerControl(optimizer = 'bobyqa'),
                  family = binomial, data = fert2)
summary(glmm2way)

Anova(glmm3way)
Anova(glmm2way)
anova(glmm2way, glmm3way)
@

<<plots1,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
par(mfrow = c(1, 2))
plot(resid(glmm3way) ~ predict(glmm3way), col = variety, data = fert2)
plot(resid(glmm3way) ~ fitted(glmm3way), col = variety, data = fert2)
legend('topright', pch = 1, col = 1:5,
       legend = levels(fert2$variety))
@

<<plots2,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
plot(allEffects(glmm3way), multiline = TRUE, type = 'response',
     se=TRUE, ylim = c(0, 1), ci.style='bars', x.var = 'n.trt', rotx = 45,
     layout = c(5, 1))
@

<<nosnms,cache=TRUE>>=
table(fert2$infected==0,fert2$variety)

noSNMS3way <- glmer(cbind(infected, total) ~ variety*n.trt*status +
                    (1|block/row), subset = variety != 'SNMS',
                    control = glmerControl(optimizer = 'bobyqa'),
                    family = binomial, data = fert2)
summary(noSNMS3way)
Anova(noSNMS3way)

noSNMS2way <- glmer(cbind(infected, total) ~ variety*n.trt + variety*status +
                    n.trt*status + (1|block/row), subset = variety != 'SNMS',
                    control = glmerControl(optimizer = 'bobyqa'),
                    family = binomial, data = fert2)
summary(noSNMS2way)
Anova(noSNMS2way)

anova(noSNMS2way, noSNMS3way)
@

<<plots3,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
par(mfrow = c(1, 2))
plot(resid(noSNMS3way) ~ predict(noSNMS3way), col = variety,
     data = fert2, subset = variety!='SNMS')
plot(resid(noSNMS3way) ~ fitted(noSNMS3way), col = variety,
     data = fert2, subset = variety!='SNMS')
legend('topright', pch = 1, col = 1:5,
       legend = levels(fert2$variety))
@

\subsection{Model Refinement}

\subsection{Interpretation}

\section{Scope of inference}
\noindent

\section{Additional Comments}
\noindent While very low infection rates are, in practice, a desirable
result, they present technical and computational problems for analysis.
In similar future studies, random sampling should be used within each
plot and the sample size should be large enough that the researchers
expect at least one infected leaf to be found in each plot.

\section{References}
%\noindent Ramsey, F.L., Schafer, D.W. (2013). {\it The Statistical Sleuth: A Course in Methods of Data Analysis, Third Edition}. Boston, MA: Brooks/Cole, Cengage Learning.\\

\noindent Gelman, A. and Hill, J. (2007). {\it Data Analysis Using
Regression and Multilevel/Hierarchical Models}. New York, NY: Cambridge
University Press.\\

%\noindent Philippi, Tom. "Topic 2(b) GRTS Spatial Sampling (for Monitoring)." GRTS Spatial Sampling (for Monitoring). Web. 17 Feb. 2015. \\
%http://science.nature.nps.gov/im/datamgmt/statistics/r/advanced/grts.cfm. \\

%\noindent Barbour, Christopher, and Wright, Wilson. ``Consulting Report''. Manuscript, 2014.

%\noindent Ramsey, F.L., Schafer, D.W. (2013). {\it The Statistical Sleuth: A Course in Methods of Data Analysis, Third Edition}. Boston, MA: Brooks/Cole, Cengage Learning.\\

\noindent Sloderbeck, J., Michaud, P., Whitworth, Robert. "Wheat Pests." CurlMite. Kansas State University, 1 May 2008. Web. 18 Sept. 2015. \verb|http://entomology.k-state.edu/extension/insect-information/crop-pests/wheat/curlmite.html|.

\end{document}

### 2015-09-29 draft1leslieandkenny.Rnw
Added some simulated data and a bit about the model.

@@
 \usepackage{graphicx}
+\usepackage{amsmath}
 
@@
 
+# If status=INOC is the most interesting case, make INOC the baseline
+fert2$status2 <- relevel(fert2$status, 'INOC')
+
 # Create a row factor
@@
 fert2$n.status <- with(fert2, interaction(n.trt, status))
+@
+
+<<fakedata,echo=FALSE>>=
+# Create some fake data with no differences anywhere
+set.seed(983724)
+fert.sim1 <- data.frame(
+  'block' = c(rep('A', 30), rep('B', 30), rep('C', 30), rep('D', 30), rep('E', 30), rep('F', 30)),
+  'variety' = c(rep('Var1', 6), rep('Var2', 6), rep('Var3', 6), rep('Var4', 6), rep('Var5', 6)),
+  'n.application' = rep(c('FALL', 'FALL', 'EARLY SPRING', 'EARLY SPRING', 'LATE SPRING', 'LATE SPRING'), 30),
+  'status' = rep(c('CNTL', 'INOC'), 90),
+  'total' = rep(30, 180),
+  'infected' = sample(fert2$infected, 180, replace = TRUE))
+
+# Correctly order nitrogen application levels
+fert.sim1$n.trt <- relevel(fert.sim1$n.application, 'FALL')
+
+# If status=INOC is the most interesting case, make INOC the baseline
+fert.sim1$status2 <- relevel(fert.sim1$status, 'INOC')
+
+# Create a row factor
+fert.sim1$row <- with(fert.sim1, interaction(variety, block))
 @
@@
 He is currently working on several projects regarding the spread and development of the Wheat Streak Mosaic Virus (WSMV) in Winter Wheat.
-Nar has taken Statistics $511$ and $512$ here at Montana State University, and he is currently enrolled in Mixed Models.
+Nar has taken Statistics $511$ and $512$ here at Montana State University, and he is currently auditing Statistics $448$.
 Nar has asked for our help in building a model, fitting it in R, and interpreting the results.
@@
 The data are collected for the first year, and the second year plants are waiting to be analyzed in the lab {\bf describe how they are stored during waiting time}.
-Planting for the third year will begin this fall {\bf I think?}.
-\\
+Planting for the third year will begin this fall {\bf I think?}.\\
 
@@
 
+Because the response variable is a binomial count, and because the design
+includes random assignment to rows within blocks, a binomial generalized
+linear mixed model is appropriate. This model will estimate the probability
+of infection for each \verb|variety:treatment:status| combination while
+controlling for differences between blocks and between
+rows in the same block.
+
+The model is
+$$
+y_i\sim\mathrm{Binomial}(30, p_i),
+$$
+$$
+\mathrm{logit}(p_i)=\mathbf{x}_i\mathbf{\beta}+b_{0,j[i]}+b_{1,j[i],k[i]}
+$$
+where
+\begin{itemize}
+\item \(p_i\) is the probability of infection for the \(i\)th observation,
+\item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
+natural logarithm of the odds of infection,
+\item \(\mathbf{x_i}\) is a row vector containing indicator variables for
+the variety, nitrogen application timing, and inoculation status of
+observation \(i\), and
+\item \(b_{0,j}\sim\mathrm{N}(0,\sigma^2_{b_0})\) and
+\(b_{1,j,k}\sim\mathrm{N}(0,\sigma^2_{b_1})\) are random effects for block
+and row, respectively.
+\end{itemize}
+
 \subsection{Graphical Data Exploration}
@@
 qplot(x = n.trt, y = infected, geom = 'boxplot', color = status,
-      facets = .~variety, data = fert2) +
+      facets = .~variety, data = fert.sim1) +
   theme(axis.text.x = element_text(angle = 90))
@@
 # Order the dataset
-#fert.ordered <- arrange(fert2, block, variety, status, n.trt)
-fert.ordered <- arrange(fert2, variety, block, status, n.trt)
+fert.ordered <- arrange(fert.sim1, variety, block, status, n.trt)
 
@@
 segments(x0 = 1.5:4.5, y0 = 0.5, y1 = 6.5, col = 'white')
-axis(3, labels = rep(levels(fert2$block), 5),
+axis(3, labels = rep(levels(fert.ordered$block), 5),
      at = seq(7/12, 5 + 5/12, 1/6), cex.axis = 0.75)
-axis(2, labels = levels(fert2$n.status), at = 1:6, las = 2)
-axis(1, labels = levels(fert2$variety), at = 1:5)
+axis(2, labels = levels(fert.ordered$n.status), at = 1:6, las = 2)
+axis(1, labels = levels(fert.ordered$variety), at = 1:5)
 
@@
                   control = glmerControl(optimizer = 'bobyqa'),
-                  family = binomial, data = fert2)
+                  family = binomial, data = fert.sim1)
 summary(glmm3way)
@@
                     control = glmerControl(optimizer = 'bobyqa'),
-                  family = binomial, data = fert2)
+                  family = binomial, data = fert.sim1)
 summary(glmm2way)
@@
 par(mfrow = c(1, 2))
-plot(resid(glmm3way) ~ predict(glmm3way), col = variety, data = fert2)
-plot(resid(glmm3way) ~ fitted(glmm3way), col = variety, data = fert2)
+plot(resid(glmm3way) ~ predict(glmm3way), col = variety, data = fert.sim1)
+plot(resid(glmm3way) ~ fitted(glmm3way), col = variety, data = fert.sim1)
 legend('topright', pch = 1, col = 1:5,
-       legend = levels(fert2$variety))
+       legend = levels(fert.sim1$variety))
 @

### 2015-09-30 draft1leslieandkenny.Rnw
Some code tweaks. (Simulated data using NBin.)

@@
 
-# Combine nitrogen and inoc/cntl
+# Combine nitrogen and inoc/cntl (for the heatmap labels)
 fert2$n.status <- with(fert2, interaction(n.trt, status))
@@
   'total' = rep(30, 180),
-  'infected' = sample(fert2$infected, 180, replace = TRUE))
+  'infected' = rnbinom(180, 4, 7/9))
+#  'infected' = sample(fert2$infected, 180, replace = TRUE))
 
@@
 fert.sim1$row <- with(fert.sim1, interaction(variety, block))
+
+# Combine nitrogen and inoc/cntl
+fert.sim1$n.status <- with(fert.sim1, interaction(n.trt, status))
 @
@@
 the variety, nitrogen application timing, and inoculation status of
-observation \(i\), and
+observation \(i\),
+\item \(\mathbf{\beta}\) is a column vector of fixed-effect coefficients, and
 \item \(b_{0,j}\sim\mathrm{N}(0,\sigma^2_{b_0})\) and
@@
 <<models,cache=TRUE>>=
+glmm3way.sim <- glmer(cbind(infected, total) ~ variety*n.trt*status +
+                        (1|block/row),
+                      control = glmerControl(optimizer = 'bobyqa',
+                                             optCtrl = list(maxfun = 20000)),
+                      family = binomial, data = fert.sim1)
+summary(glmm3way.sim)
+Anova(glmm3way.sim)
+
 glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status +
@@
                   control = glmerControl(optimizer = 'bobyqa'),
-                  family = binomial, data = fert.sim1)
+                  family = binomial, data = fert2)
 summary(glmm3way)
@@
                     control = glmerControl(optimizer = 'bobyqa'),
-                  family = binomial, data = fert.sim1)
+                  family = binomial, data = fert2)
 summary(glmm2way)

### 2015-09-30 draft1leslieandkenny.Rnw
Made some progress on model and graphics.

--- 2015-09-23 draft1leslieandkenny.Rnw
@@
 require(tabplot)
-require(dplyr)
 require(car)
 require(lme4)
-require(ggplot2)
 require(lattice)
-require(effects)
 @
@@
 
-# Combine nitrogen and inoc/cntl
+# Combine nitrogen and inoc/cntl (for the heatmap labels)
 fert2$n.status <- with(fert2, interaction(n.trt, status))
+@
+
+<<fakedata,echo=FALSE>>=
+# Create some fake data with no differences anywhere
+set.seed(983724)
+fert.sim1 <- data.frame(
+  'block' = c(rep('A', 30), rep('B', 30), rep('C', 30), rep('D', 30), rep('E', 30), rep('F', 30)),
+  'variety' = c(rep('Var1', 6), rep('Var2', 6), rep('Var3', 6), rep('Var4', 6), rep('Var5', 6)),
+  'n.application' = rep(c('FALL', 'FALL', 'EARLY SPRING', 'EARLY SPRING', 'LATE SPRING', 'LATE SPRING'), 30),
+  'status' = rep(c('CNTL', 'INOC'), 90),
+  'total' = rep(30, 180),
+  'infected' = rnbinom(180, 4, 7/9))
+#  'infected' = sample(fert2$infected, 180, replace = TRUE))
+@
+
+<<fakepre,echo=FALSE>>=
+# Create a row factor
+fert.sim1$row <- with(fert.sim1, interaction(variety, block))
+
+# Correctly order nitrogen application levels, the default is alphabetical
+fert.sim1$n.trt <- relevel(fert.sim1$n.application, 'FALL')
+
+# If status=INOC is the most interesting case, make INOC the baseline
+fert.sim1$status2 <- relevel(fert.sim1$status, 'INOC')
 @
@@
 
+\noindent Because the response variable is a Binomial count, and because
+the design includes random assignment of variety to rows within blocks,
+a binomial generalized linear mixed model is appropriate. This model
+will estimate the probability of infection for each
+\verb|variety:treatment:status| combination while controlling for
+differences between blocks and between rows in the same block.
+
+\noindent The model is
+$$
+y_i\sim\mathrm{Binomial}(30, p_i),
+$$
+$$
+\mathrm{logit}(p_i)=\mathbf{x}_i\mathbf{\beta}+b_{0,j[i]}+b_{1,j[i],k[i]}
+$$
+where
+\begin{itemize}
+\item \(p_i\) is the probability of infection in the \(i\)th plot,
+\item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
+natural logarithm of the odds of infection,
+\item \(\mathbf{x_i}\) is a row vector containing indicator variables for
+the variety, nitrogen application timing, and inoculation status of
+observation \(i\),
+\item \(\mathbf{\beta}\) is a column vector of fixed-effect coefficients, and
+\item \(b_{0,j}\sim\mathrm{N}(0,\sigma^2_{b_0})\) and
+\(b_{1,j,k}\sim\mathrm{N}(0,\sigma^2_{b_1})\) are random effects for block
+and row, respectively.
+\end{itemize}
+
+\noindent The assumptions for this model are that plants within a plot
+are independent and have the same probability of being infected, and
+that plots within a row are endependent of each other.
+
+\noindent Advantages of the Binomial GLMM include:
+\begin{itemize}
+\item It estimates and controls for variability between rows and blocks.
+\item It avoids confounding variety effects with row effects and correctly
+models the rows and nested within blocks.
+\item Random effects can explain extra variation beyond what the
+fixed-effect Binomial GLM can account for.
+\end{itemize}
+
+\noindent Disadvantages of this model include:
+\begin{itemize}
+\item It must be fit by numerically maximizing the likelihood with an
+iterative algorithm. Including a large number of parameters (such as
+third-order interaction coefficients) can cause the algorithm to fail
+to converge.
+\item The domain of the logit function is \((0, 1)\). If a large proportion
+of the data are zeros, meaning an infection probability is near 0, the
+algorithm will have difficulty estimating the coefficients. In this
+situation, estimates may be unreliable or the algorithm may not converge.
+\item We are not aware of any tools that can fit a Binomial GLMM with
+overdispersion. If there are additional sources of variation that the
+model does not account for, the standard error estimates will be too small.
+\end{itemize}
+
 \subsection{Graphical Data Exploration}
 
-<<interactionplot,echo=FALSE>>=
+\noindent Before fitting a model, we recommend starting by creating plots
+for exploratory data analysis. This is useful for seeing which interactions
+may be present, and for identifying patterns that may represent violations
+of assumptions or potential computational problems.
+
+\noindent A good way to visually compare varieties, treatments, and
+inoculation status is with boxplots of infection counts for each
+combination of factor levels. The \verb|qplot| function in the
+\verb|ggplot2| package is a simple way to do this.
+
+\begin{figure}
+\caption{Example interaction plot using \texttt{qplot}.}
+<<interactionplot,echo=FALSE,message=FALSE>>=
+require(ggplot2)
 qplot(x = n.trt, y = infected, geom = 'boxplot', color = status,
-      facets = .~variety, data = fert2) +
+      facets = .~variety, data = fert.sim1) +
   theme(axis.text.x = element_text(angle = 90))
 @
+\end{figure}
 
-<<heatmap,echo=FALSE>>=
+\noindent We noticed that infection counts of 0 were recored in 67 of the
+180 plots in Nar's dataset. This is a large fraction of the data; if the
+distribution of these zeros is related to the variables in the model then
+it may be impossible to estimate coefficients for factor levels where
+infection rates are nearly zero.
+
+\noindent We recommend sorting the data and creating a heatmap to look
+for patterns of zeros. Figure 2 shows an example using simulated data.
+The most serious problems will come from the zeros, so we chose a contrasting
+color (black) for plots with infection counts of 0.
+
+\begin{figure}
+\caption{Example heatmap visualization for identifying patterns of zeros.}
+<<heatmap,echo=FALSE,message=FALSE>>=
 ## Heatmap with meaningful ordering
 
-# Order the dataset
-#fert.ordered <- arrange(fert2, block, variety, status, n.trt)
-fert.ordered <- arrange(fert2, variety, block, status, n.trt)
+# Order the dataset, using the arrange() function from the dplyr package
+require(dplyr)
+fert.ordered <- arrange(fert.sim1, variety, block, status, n.trt)
 
@@
 # Plot the heatmap, with zeros in black and segments separating the blocks
-par(mar = c(5, 10, 6, 2))
-image(x = seq(0.5, 5.5, 1/6), y = 1:6, z = infected.arranged,
-      col = c('black', heat.colors(30)), xlab = '', ylab = '',
-      zlim = c(0, 30), yaxt = 'n', xaxt = 'n')
+par(mar = c(5, 10, 6, 2)) # Set big margins
+image(z = infected.arranged, y = 1:6,
+
+      # Variety blocks are 1 unit wide, centered at 0.5, 1.5, etc
+      x = seq(0.5, 5.5, 1/6),
+
+      # Use black for 0, and use heatmap colors (red-orange-yellow-white) for 1 to 30
+      col = c('black', heat.colors(30)), zlim = c(0, 30),
+      xlab = '', ylab = '', yaxt = 'n', xaxt = 'n') # Don't automatically create axes or labels
+
+# Use white line segments to visually separate the varieties
+segments(x0 = 1.5:4.5, y0 = 0.5, y1 = 6.5, col = 'white')
+
+# Place a title at the top, and label Varieties, treatment:status levels, and blocks around the image
 title('Infection Counts', line = 4)
-segments(x0 = 1.5:4.5, y0 = 0.5, y1 = 6.5, col = 'white')
-axis(3, labels = rep(levels(fert2$block), 5),
-     at = seq(7/12, 5 + 5/12, 1/6), cex.axis = 0.75)
-axis(2, labels = levels(fert2$n.status), at = 1:6, las = 2)
-axis(1, labels = levels(fert2$variety), at = 1:5)
+axis(3, labels = rep(levels(fert.ordered$block), 5), cex.axis = 0.75,
+
+     # Each block is plotted in a column with width 1/6, so put the labels in the middle
+     at = seq(7/12, 5 + 5/12, 1/6))
+axis(2, labels = levels(with(fert.ordered, interaction(n.trt, status))), at = 1:6, las = 2)
+axis(1, labels = levels(fert.ordered$variety), at = 1:5)
 
@@
 par(mar = c(5, 1, 6, 2))
-image(y = -0.5:30.5, z = matrix(0:30, nrow = 1), axes = FALSE, ylab = '',
+image(y = seq(-0.5, 30.5, 1), z = matrix(0:30, nrow = 1), axes = FALSE, ylab = '',
       col = c('#000000', heat.colors(30)), zlim = c(0, 30))
@@
 @
+\end{figure}
 
+\noindent If information about the physical layout of the field is
+available, it would be possible to construct a similar heatmap to look
+for spatial patterns. If correlation between nearby plots is seen,
+this should be included in the model.
+
+\noindent Another option for examining the data graphically is the
+\verb|itableplot| function in the \verb|tabplot| package. This is a
+useful interactive tool for visualizing large datasets and identifying
+patterns.
+
+\begin{figure}
+\caption{Example \texttt{itableplot} output.}
 <<tabplot,echo=FALSE,warning=FALSE,message=FALSE>>=
@@
 @
+\end{figure}
 
-The \verb|itableplot| function in the \verb|tabplot| package is another
-useful tool for visualizing the data and identifying patterns.
+\subsection{Model Fitting and Possible Issues}
+\noindent The
+<<models1,warning=FALSE,results='hide',cache=TRUE>>=
+glmm3way.sim <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
+                      control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
+                      family = binomial, data = fert.sim1)
+@
 
-\subsection{Convergeance Warning Messages}
+<<models2,warning=FALSE,results='hide',cache=TRUE>>=
+noVar1.sim <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
+                      control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
+                      family = binomial, data = fert.sim1, subset = variety!='Var1')
+@
 
-<<models,cache=TRUE>>=
-glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status +
-                    (1|block/row),
+<<models3,warning=FALSE,results='hide',cache=TRUE>>=
+glmm3way.sim <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
+                      control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
+                      family = binomial, data = fert.sim1)
+@
+
+<<models4,echo=FALSE,warning=FALSE,results='hide',cache=TRUE>>=
+glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
                   control = glmerControl(optimizer = 'bobyqa'),
                   family = binomial, data = fert2)
-summary(glmm3way)
+#summary(glmm3way)
 
-glmm2way <- glmer(cbind(infected, total) ~ variety*n.trt + variety*status +
-                    n.trt*status + (1|block/row),
+glmm2way <- glmer(cbind(infected, total) ~ variety*n.trt + variety*status2 +
+                    n.trt*status2 + (1|block/row),
                     control = glmerControl(optimizer = 'bobyqa'),
                   family = binomial, data = fert2)
-summary(glmm2way)
+#summary(glmm2way)
 
-Anova(glmm3way)
-Anova(glmm2way)
-anova(glmm2way, glmm3way)
+#Anova(glmm3way)
+#Anova(glmm2way)
+#anova(glmm2way, glmm3way)
 @
 
-<<plots1,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
+<<plots1,eval=FALSE,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
 par(mfrow = c(1, 2))
-plot(resid(glmm3way) ~ predict(glmm3way), col = variety, data = fert2)
-plot(resid(glmm3way) ~ fitted(glmm3way), col = variety, data = fert2)
+plot(resid(glmm3way) ~ predict(glmm3way), col = variety, data = fert.sim1)
+plot(resid(glmm3way) ~ fitted(glmm3way), col = variety, data = fert.sim1)
 legend('topright', pch = 1, col = 1:5,
-       legend = levels(fert2$variety))
-@
-
-<<plots2,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
-plot(allEffects(glmm3way), multiline = TRUE, type = 'response',
-     se=TRUE, ylim = c(0, 1), ci.style='bars', x.var = 'n.trt', rotx = 45,
-     layout = c(5, 1))
+       legend = levels(fert.sim1$variety))
 @
@@
 
-noSNMS3way <- glmer(cbind(infected, total) ~ variety*n.trt*status +
+noSNMS3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
                     (1|block/row), subset = variety != 'SNMS',
@@
 
-noSNMS2way <- glmer(cbind(infected, total) ~ variety*n.trt + variety*status +
-                    n.trt*status + (1|block/row), subset = variety != 'SNMS',
+noSNMS2way <- glmer(cbind(infected, total) ~ variety*n.trt + variety*status2 +
+                    n.trt*status2 + (1|block/row), subset = variety != 'SNMS',
                     control = glmerControl(optimizer = 'bobyqa'),
@@
 
-<<plots3,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
+<<plots3,eval=FALSE,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
 par(mfrow = c(1, 2))
@@
 @
+
+\begin{figure}
+\caption{The fitted model can be nicely summarized with plots from the \texttt{effects} package.}
+<<plots2,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
+require(effects)
+plot(allEffects(glmm3way.sim), multiline = TRUE, type = 'response', se = TRUE,
+     ci.style='bars', x.var = 'n.trt', rotx = 45, layout = c(5, 1),
+     main = 'Variety:Treatment:Inoculation Effects Plot',
+     xlab = 'Nitrogen Application Date', ylab = 'Infection Probability',
+     key.args = list(title = 'Inoculation Status'))
+@
+\end{figure}
 
@@
 expect at least one infected leaf to be found in each plot.
+
+\section{Appendix: R Code}
+\noindent Since Nar expressed a desire for assistance with R, we have
+included the code we used to create the output for this report. Note
+that we used a simulated dataset.
+
+\subsection{Data Preprocessing}
+\noindent We created variable with a unique level for each row, since row
+it was not included in the dataset. We also changed the baseline levels
+of \verb|n.application| and \verb|status|.
+<<fakepre,eval=FALSE>>=
+@
+
+\subsection{\texttt{qplot}}
+<<interactionplot,eval=FALSE>>=
+@
+
+\subsection{Heatmap}
+Nar only needs to change \verb|fert.sim1| inside the \verb|arrange| call
+to the name of his data frame. The rest of this code is self-contained and
+will create the heatmap with appropriate labels for varieties, treatments,
+and blocks.
+<<heatmap,eval=FALSE>>=
+@
+
+\subsection{Summary Effects Plot}
+<<plots2,eval=FALSE>>=
+@
 

### 2015-09-30 leslieandkenny.Rnw
Progress

--- 2015-09-30 draft1leslieandkenny.Rnw
@@
 require(lattice)
+require(nlme)
 @
@@
 # Read and set up the data
-fert2 <- read.csv('fert2.csv')
+fert2 <- read.csv("fert2.csv")
 
@@
 He is currently working on several projects regarding the spread and development of the Wheat Streak Mosaic Virus (WSMV) in Winter Wheat.
-Nar has taken Statistics $511$ and $512$ here at Montana State University, and he is currently auditing Statistics $448$.
+Nar has taken Statistics $511$ and $512$ here at Montana State University, and he is currently sitting in Mixed Models.
 Nar has asked for our help in building a model, fitting it in R, and interpreting the results.
@@
 Each row was randomly assigned to a variety of Winter Wheat with separate randomizations in each block.
-The rows then were divided into six $1.5$ by $5$ meter plots.
+Three of the chosen varieties are known to be resistant to WSMV (SNMS, TAM 112, and MACE).
+The pronghorn variety (PRHG) is known to be susceptible to WSMV.
+The last variety, Yellowstone (YLST), is a common variety in Montana, and its susceptibility to WSMV is unknown.
+\\
+
+\noindent After assigning rows to varieties, the rows were then divided into six $1.5$ by $5$ meter plots.
 Each plot was randomly assigned to one of six combinations of nitrogen application time and inoculation status, with separate randomizations in each row.
 The combinations were fall/inoculated, fall/control, early spring/inoculated, early spring/control, late spring/inoculated, and late spring/control.
-All plots were planted in the {\bf FALL?} and allowed to grow over the winter.
-The following {\bf September?}, $30$ plants were chosen from each plot to be taken to the lab and tested for presence of the virus.
-The selection of the plants from each plot was not random nor technically systematic.
-The {\bf undergraduate?} technicians were instructed to collect a sample of plants spread evenly throughout the field.
-{\bf different undergraduates doing picking?
-picking done all at the same time?
-is there visual evidence of the disease?} A total of $180$ plants were collected.
-The plants were then sent to the lab and screened for the virus via the ELISA procedure.
-This is a three year study.
-The data are collected for the first year, and the second year plants are waiting to be analyzed in the lab {\bf describe how they are stored during waiting time}.
-Planting for the third year will begin this fall {\bf I think?}.\\
+\\
 
 \noindent Plots that were assigned to the inoculated treatment had five infected plants transplanted to the middle of the plot.
-These plants were infected {\bf as adults?} in the greenhouse by clipping an infected leaf to the healthy plant.
+These plants were infected in the greenhouse by clipping an infected leaf to the healthy plant.
+The plants are infected in the tillering stage of their life cycle at an age of about one month.
 The Wheat Streak Mosaic Virus is transmitted by the wheat curl mite, tiny organisms that are nearly invisible to the naked eye.
 They can easily move from leaf to leaf on a plant, and they are transported from plant to plant by the wind  (Sloderbeck $2008$).
-It is our understanding that once the infected mites were introduced to the study area, they were considered to be everywhere in the air and that all plants in the field were exposed to the infected mites.
+It is conceivable that the mites on the infected plants introduced to the inoculated plots could move to healthy plants within that plot.
+But, the researchers say that it is nearly impossible for the mites to move from an inoculated plot to a non-inoculated plot with such a low population of mites.
+Plants with the virus in non-inoculated plots are assumed to have been infected by mites in the surrounding environment.
+\\
+
+\noindent All plots were planted in mid September and allowed to grow over the winter.
+The following spring and summer, $30$ leaves were picked from different plants within each plot and taken to the lab to be tested for presence of the virus.
+The selection of the leaves from each plot was not random nor technically systematic.
+The six technicians were instructed to collect a sample of plants spread evenly throughout the plot.
+There is no possibility of visual bias because the symptoms are not heavy enough in this area of Montana for the plants to show clear visual evidence of the virus.
+\\
+
+\noindent The sampling was all done in one day on two occasions.
+The first sampling date occured at the end of May when the plants were in the tillering stage, and the second sampling date occured in early July when plants were in the flowering stage.
+In the first year of the study, $2013-2014$, all plots were sampled on both dates.
+In the second year of the study, only plots with susceptible varieties were sampled on the first date, and all plots were sampled on the second date The advice given in this report is only for analysis of the data collected on the second date in July.
+After the leaves were picked in the field, they were sent to the lab and screened for the virus via the ELISA procedure.
+The data are collected and organized in a spreadsheet for the first year.
+The second year leaves are currently stored in ziploc bags in the freezer and are waiting to be analyzed in the lab.
+The researchers do not expect this waiting period to affect their ability to detect the virus.
+\\
+
+\noindent This is a three year study, and planting for the third year of the study is currently taking place.
+Each year, the assignment of varieties to rows and treatments to plots is re-randomized.
+Furthermore, planting takes place in the gaps between the plots from the previous year.
 
@@
 are independent and have the same probability of being infected, and
-that plots within a row are endependent of each other.
+that plots within a row are independent of each other.
 
@@
 
-\begin{figure}
+\begin{figure}[b]
 \caption{Example interaction plot using \texttt{qplot}.}
@@
 
-\begin{figure}
+\noindent Any treatment or variety with many black squares is a reason
+for concern. In the example, all but one of the inoculated fall-application
+plots for Var1 had counts of 0. The software will have difficulty estimating
+the the three-way interaction involving fall nitrogen application and
+variety 1.
+
+\noindent If a treatment has many zeros across all varieties or a variety
+has many zeros, Nar should consider whether he has reason to believe
+that treatment or variety has qualitative differences from the others
+that would cause it to have a much lower infrection rate. \textbf{If and
+only if it is justified}, Nar could omit all of the data for this treatment
+or variety when fitting the model. The scope of inference for this model
+must then be limited to varieties and treatments that have a nonzero
+infection rate.
+
+\begin{figure}[b]
 \caption{Example heatmap visualization for identifying patterns of zeros.}
@@
 for spatial patterns. If correlation between nearby plots is seen,
-this should be included in the model.
+the model should be modyfied to account for this.
 
@@
 
-\begin{figure}
+\begin{figure}[b]
 \caption{Example \texttt{itableplot} output.}
@@
 \subsection{Model Fitting and Possible Issues}
-\noindent The
+
+The following code will fit the three-way interaction model:
+
 <<models1,warning=FALSE,results='hide',cache=TRUE>>=
-glmm3way.sim <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
+glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
                   control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
@@
 
+We noticed that the SNMS variety had 27 out of 36 counts that were zeros.
+Code like the following
+
 <<models2,warning=FALSE,results='hide',cache=TRUE>>=
-noVar1.sim <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
+noVar1 <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
                 control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
@@
 
+If there is no interest in comparing between inoculated plots and
+control plots, separate models could be fit.
+
 <<models3,warning=FALSE,results='hide',cache=TRUE>>=
-glmm3way.sim <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
+onlyINOC <- glmer(cbind(infected, total) ~ variety*n.trt + (1|block/row),
                   control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
-                      family = binomial, data = fert.sim1)
+                  family = binomial, data = fert.sim1, subset = status=='INOC')
+
+onlyCNTL <- glmer(cbind(infected, total) ~ variety*n.trt + (1|block/row),
+                  control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
+                  family = binomial, data = fert.sim1, subset = status=='CNTL')
 @
 
-<<models4,echo=FALSE,warning=FALSE,results='hide',cache=TRUE>>=
+<<models5,echo=FALSE,warning=FALSE,results='hide',cache=TRUE>>=
 glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
@@
 
-<<nosnms,cache=TRUE>>=
+<<nosnms,echo=FALSE,warning=FALSE,results='hide',cache=FALSE>>=
 table(fert2$infected==0,fert2$variety)
@@
 
+<<glmmodel1,cache=FALSE>>=
+glm1 <- glm(cbind(infected, total) ~ variety*n.trt*status2 + row,
+            family = quasibinomial, data = fert.sim1)
+glm2 <- glm(cbind(infected, total) ~ block + variety*n.trt*status2,
+            family = quasibinomial, data = fert.sim1)
+glm3 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
+            family = quasibinomial, data = fert.sim1)
+glm4 <- glm(cbind(infected, total) ~ block/row + variety*n.trt*status2,
+            family = quasibinomial, data = fert.sim1)
+glm5 <- glm(cbind(infected, total) ~ block/variety + variety*n.trt*status2,
+            family = quasibinomial, data = fert.sim1)
+
+glmnointMF <- glm(cbind(infected, total) ~ block/variety + variety*n.trt*status2,
+            family = quasibinomial, data = fert2)
+glmmnointMF <- glmer(cbind(infected, total) ~ variety + variety*n.trt*status2 + (1|block/row),
+            family = binomial, data = fert.sim1)
+@
+
 \subsection{Model Refinement}
+
+Recommend to use Chi Squared tests or F-tests? I think F-tests but not exactly sure what the F-test is giving or where its coming from.
+
+Correct df? different from lme and glmer...
+
+<<nosnms3way, echo=FALSE>>=
+Anova(noSNMS3way)
+anova(noSNMS3way)
+#1-pf(2.29, 6, 118)
+
+AIC(noSNMS3way, noSNMS2way)
+anova(noSNMS3way, noSNMS2way)
+
+#lme model to get correct df for f tests
+#why is den df now 100?
+require(nlme, quietly=T)
+noSNMS3waylme <- lme(infected ~ variety*n.trt*status,
+                    random=~1|block/row, method="ML",
+                     data = subset(fert2, variety != "SNMS"))
+
+glm3way <- glm(cbind(infected, total) ~ variety*n.trt*status+row, family="binomial", data=subset(fert2, variety != "SNMS"))
+@
+
+explain: if no evidence of a three way interaction what can you do, how to continue selection.
 
@@
 
+Explain how to interpret with three way interaction (essentially interpret two way interaction for each inoculated and control)
+
 \section{Scope of inference}
-\noindent
+
+The field where the study took place was not randomly selected from a larger population of fields, so inference from this study does not extend beyond the field at the base of the Bridger mountains where the experiment was conducted.
+There are potential confounding factors in this field that may not be present in other fields around Montana.
+For example, the soil in this specific field may have high starting levels of nitrogen from previous studies that could affect the observed relationship between nitrogen application time and probability of virus infection.
+If inference beyond the study field is desired, it would be necessary to justify that the study field is similar to other wheat fields in the population of interest in terms of soil composition and all other variables.
+\\
+
+\noindent Plants were randomly assigned to treatments, so it is justified to infer that the observed infection rates in this field are caused by the timing of the nitrogen application and the inoculation status of the plots.
+A causal relationship {\it cannot} be inferred between variety and virus infection, however, because plants were not randomly assigned to varieties.
 
@@
 
-%\noindent Ramsey, F.L., Schafer, D.W. (2013). {\it The Statistical Sleuth: A Course in Methods of Data Analysis, Third Edition}. Boston, MA: Brooks/Cole, Cengage Learning.\\
+\noindent Ramsey, F.L., Schafer, D.W. (2013). {\it The Statistical Sleuth: A Course in Methods of Data Analysis, Third Edition}. Boston, MA: Brooks/Cole, Cengage Learning.\\
 
-\noindent Sloderbeck, J., Michaud, P., Whitworth, Robert.
-"Wheat Pests." CurlMite.
-Kansas State University, 1 May 2008.
-Web.
-18 Sept.
-2015.
+\noindent Sloderbeck, J., Michaud, P., Whitworth, Robert. "Wheat Pests." CurlMite. Kansas State University, 1 May 2008. Web. 18 Sept. 2015.\\
 \verb|http://entomology.k-state.edu/extension/insect-information/crop-pests/wheat/curlmite.html|.

### 2015-09-30 draft1leslieandkenny.Rnw
Changed code to glm. 3rd try.

@@
 
+# If YSTN is the variety we want to know about the most, make it baseline
+fert2$variety2 <- relevel(fert2$variety, 'YSTN')
+
 # Create a row factor
@@
 # Combine nitrogen and inoc/cntl (for the heatmap labels)
-fert2$n.status <- with(fert2, interaction(n.trt, status))
+#fert2$n.status <- with(fert2, interaction(n.trt, status))
 @
@@
 <<fakepre,echo=FALSE>>=
-# Create a row factor
-fert.sim1$row <- with(fert.sim1, interaction(variety, block))
-
 # Correctly order nitrogen application levels, the default is alphabetical
@@
 
-\begin{figure}
+\noindent Any treatment or variety with many black squares is a reason
+for concern. In the example, all but one of the inoculated fall-application
+plots for Var1 had counts of 0. The software will have difficulty estimating
+the the three-way interaction involving fall nitrogen application and
+variety 1.
+
+\noindent If a treatment has many zeros across all varieties or a variety
+has many zeros, Nar should consider whether he has reason to believe
+that treatment or variety has qualitative differences from the others
+that would cause it to have a much lower infrection rate. These data would
+be considered as coming from a different population than the data with
+fewer zeros. \textbf{If and only if it is justified}, Nar could omit all
+of the data for this treatment or variety when fitting the model. The scope
+of inference for this model must then be limited to varieties and
+treatments that have a nonzero infection rate. A separate analysis would
+be needed for more resistant population.
+
+\begin{figure}[b]
 \caption{Example heatmap visualization for identifying patterns of zeros.}
@@
 
-# Order the dataset, using the arrange() function from the dplyr package
+# Order the dataset, using the arrange function from the dplyr package
 require(dplyr)
@@
       col = c('black', heat.colors(30)), zlim = c(0, 30),
-      xlab = '', ylab = '', yaxt = 'n', xaxt = 'n') # Don't automatically create axes or labels
+
+      # Don't automatically create axes or labels
+      xlab = '', ylab = '', yaxt = 'n', xaxt = 'n')
 
@@
 for spatial patterns. If correlation between nearby plots is seen,
-this should be included in the model.
+the model should be modified to account for this.
 
@@
 <<tabplot,echo=FALSE,warning=FALSE,message=FALSE>>=
-tableplot(fert.ordered, select = c(infected, variety, block, n.trt, status),
+tableplot(fert.sim1, select = c(infected, variety, block, n.trt, status),
           nBins = 30, sortCol = variety)
@@
 \subsection{Model Fitting and Possible Issues}
-\noindent The
-<<models1,warning=FALSE,results='hide',cache=TRUE>>=
-glmm3way.sim <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
-                      control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
-                      family = binomial, data = fert.sim1)
+
+The following code will fit the three-way interaction model:
+
+<<startingmodel,warning=FALSE,results='hide',cache=TRUE>>=
+glm3way <- glm(cbind(infected, total) ~ block/variety + variety*n.trt*status2,
+               family = quasibinomial, data = fert.sim1)
 @
 
-<<models2,warning=FALSE,results='hide',cache=TRUE>>=
-noVar1.sim <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
-                      control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
-                      family = binomial, data = fert.sim1, subset = variety!='Var1')
+If it is decided that certain observations come from an extremely
+resistant population and can be omited from the current analysis.
+
+<<novar1,warning=FALSE,results='hide',cache=TRUE>>=
+noVar1 <- glm(cbind(infected, total) ~ block/variety + variety*n.trt*status2,
+              family = quasibinomial, data = fert.sim1, subset = variety!='Var1')
 @
 
-<<models3,warning=FALSE,results='hide',cache=TRUE>>=
-glmm3way.sim <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
-                      control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
-                      family = binomial, data = fert.sim1)
+If there is no interest in comparing between inoculated plots and
+control plots, separate models could be fit.
+
+<<separatemodels,warning=FALSE,results='hide',cache=TRUE>>=
+onlyINOC <- glm(cbind(infected, total) ~ block/variety + variety*n.trt,
+                family = quasibinomial, data = fert.sim1, subset = status=='INOC')
+
+onlyCNTL <- glm(cbind(infected, total) ~ block/variety + variety*n.trt,
+                family = quasibinomial, data = fert.sim1, subset = status=='CNTL')
 @
 
-<<models4,echo=FALSE,warning=FALSE,results='hide',cache=TRUE>>=
-glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
-                  control = glmerControl(optimizer = 'bobyqa'),
+<<models5,echo=FALSE,warning=FALSE,results='hide',cache=TRUE>>=
+glm3way2 <- glm(cbind(infected, total) ~ block/variety + variety*n.trt*status2,
                 family = binomial, data = fert2)
-#summary(glmm3way)
+summary(glm3way2)
 
-glmm2way <- glmer(cbind(infected, total) ~ variety*n.trt + variety*status2 +
-                    n.trt*status2 + (1|block/row),
-                    control = glmerControl(optimizer = 'bobyqa'),
-                  family = binomial, data = fert2)
-#summary(glmm2way)
+glm2way2 <- glm(cbind(infected, total) ~ block/variety + variety*n.trt + variety*status2 + n.trt*status2,
+                family = quasibinomial, data = fert2)
+summary(glm2way2)
 
-#Anova(glmm3way)
-#Anova(glmm2way)
-#anova(glmm2way, glmm3way)
+Anova(glm3way2, test.statistic = 'F')
+Anova(glm2way2, test.statistic = 'F')
+Anova(glm2way2, glm3way2, test.statistic = 'F')
+anova(glm2way2, glm3way2)
 @
 
-<<plots1,eval=FALSE,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
-par(mfrow = c(1, 2))
-plot(resid(glmm3way) ~ predict(glmm3way), col = variety, data = fert.sim1)
-plot(resid(glmm3way) ~ fitted(glmm3way), col = variety, data = fert.sim1)
-legend('topright', pch = 1, col = 1:5,
-       legend = levels(fert.sim1$variety))
-@
-
-<<nosnms,cache=TRUE>>=
+<<nosnms,echo=FALSE,warning=FALSE,results='hide',cache=FALSE>>=
 table(fert2$infected==0,fert2$variety)
@@
 require(effects)
-plot(allEffects(glmm3way.sim), multiline = TRUE, type = 'response', se = TRUE,
-     ci.style='bars', x.var = 'n.trt', rotx = 45, layout = c(5, 1),
-     main = 'Variety:Treatment:Inoculation Effects Plot',
+plot(Effect(c('variety','n.trt','status2'), glm3way),
+     multiline = TRUE, type = 'response', se = TRUE, ci.style='bars',
+     x.var = 'n.trt', rotx = 45, layout = c(5, 1),
+     main = 'Variety/Treatment/Inoculation Effects Plot',
      xlab = 'Nitrogen Application Date', ylab = 'Infection Probability',
@@
 
+Recommend to use Chi Squared tests or F-tests? I think F-tests but not exactly sure what the F-test is giving or where its coming from.
+
+Correct df? different from lme and glmer...
+
+<<nosnms3way, echo=FALSE>>=
+Anova(noSNMS3way)
+anova(noSNMS3way)
+#1-pf(2.29, 6, 118)
+
+AIC(noSNMS3way, noSNMS2way)
+anova(noSNMS3way, noSNMS2way)
+
+#lme model to get correct df for f tests
+#why is den df now 100?
+require(nlme, quietly=T)
+noSNMS3waylme <- lme(infected ~ variety*n.trt*status,
+                    random=~1|block/row, method="ML",
+                     data = subset(fert2, variety != "SNMS"))
+
+glm3way <- glm(cbind(infected, total) ~ variety*n.trt*status+row, family="binomial", data=subset(fert2, variety != "SNMS"))
+@
+
+explain: if no evidence of a three way interaction what can you do, how to continue selection.
+
 \subsection{Interpretation}
 
+Explain how to interpret with three way interaction (essentially interpret two way interaction for each inoculated and control)
+
 \section{Scope of inference}
-\noindent
+
+The field where the study took place was not randomly selected from a larger population of fields, so inference from this study does not extend beyond the field at the base of the Bridger mountains where the experiment was conducted.
+There are potential confounding factors in this field that may not be present in other fields around Montana.
+For example, the soil in this specific field may have high starting levels of nitrogen from previous studies that could affect the observed relationship between nitrogen application time and probability of virus infection.
+If inference beyond the study field is desired, it would be necessary to justify that the study field is similar to other wheat fields in the population of interest in terms of soil composition and all other variables.
+\\
+
+\noindent Plants were randomly assigned to treatments, so it is justified to infer that the observed infection rates in this field are caused by the timing of the nitrogen application and the inoculation status of the plots.
+A causal relationship {\it cannot} be inferred between variety and virus infection, however, because plants were not randomly assigned to varieties.
 
@@
 \subsection{Data Preprocessing}
-\noindent We created variable with a unique level for each row, since row
-it was not included in the dataset. We also changed the baseline levels
-of \verb|n.application| and \verb|status|.
+We changed the baseline levels of \verb|n.application| and \verb|status|.
+If the YLST variety is the primary variety of interest, the \verb|relevel|
+function could be used on the \verb|variety| variable the same way.
 <<fakepre,eval=FALSE>>=

### 2015-10-01 draft1leslieandkenny.Rnw
Wrote out glm model and ran spellcheck.

@@
 The pronghorn variety (PRHG) is known to be susceptible to WSMV.
-The last variety, Yellowstone (YLST), is a common variety in Montana, and its susceptibility to WSMV is unknown.
+The last variety, Yellowstone (YSTN), is a common variety in Montana, and its susceptibility to WSMV is unknown.
 \\
@@
 \noindent The sampling was all done in one day on two occasions.
-The first sampling date occured at the end of May when the plants were in the tillering stage, and the second sampling date occured in early July when plants were in the flowering stage.
+The first sampling date occurred at the end of May when the plants were in the tillering stage, and the second sampling date occurred in early July when plants were in the flowering stage.
 In the first year of the study, $2013-2014$, all plots were sampled on both dates.
@@
 The data are collected and organized in a spreadsheet for the first year.
-The second year leaves are currently stored in ziploc bags in the freezer and are waiting to be analyzed in the lab.
+The second year leaves are currently stored in Ziploc bags in the freezer and are waiting to be analyzed in the lab.
 The researchers do not expect this waiting period to affect their ability to detect the virus.
@@
 
-\subsection{Binomial Generalized Linear Mixed Model}
+\subsection{Binomial Generalized Linear Model}
 
@@
 the design includes random assignment of variety to rows within blocks,
-a binomial generalized linear mixed model is appropriate. This model
-will estimate the probability of infection for each
-\verb|variety:treatment:status| combination while controlling for
-differences between blocks and between rows in the same block.
+a Binomial generalized linear model is appropriate. We recommend using
+a three-way interaction to estimate the probability of infection for each
+variety/treatment/inoculation status combination and a block/variety
+interaction to control for differences between blocks and between rows
+in the same block.
 
@@
 $$
-\mathrm{logit}(p_i)=\mathbf{x}_i\mathbf{\beta}+b_{0,j[i]}+b_{1,j[i],k[i]}
+\mathrm{logit}(p_i)=\mu+\alpha_{b[i]}+\beta_{v[i]}+\gamma_{t[i]}+\delta_{j[i]}
++(\alpha\beta)_{b[i],v[i]}+(\beta\gamma)_{v[i],t[i]}+(\beta\delta)_{v[i],j[i]}
++(\gamma\delta)_{t[i],j[i]}+(\beta\gamma\delta)_{v[i],t[i],j[i]}
 $$
@@
 \item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
-natural logarithm of the odds of infection,
-\item \(\mathbf{x_i}\) is a row vector containing indicator variables for
-the variety, nitrogen application timing, and inoculation status of
-observation \(i\),
-\item \(\mathbf{\beta}\) is a column vector of fixed-effect coefficients, and
-\item \(b_{0,j}\sim\mathrm{N}(0,\sigma^2_{b_0})\) and
-\(b_{1,j,k}\sim\mathrm{N}(0,\sigma^2_{b_1})\) are random effects for block
-and row, respectively.
+natural logarithm of the odds of infection, and
+\item \(b[i]\), \(v[i]\), \(t[i]\), and \(j[i]\) index the block,
+variety, nitrogen application timing, and inoculation status of
+observation \(i\).
 \end{itemize}
@@
 are independent and have the same probability of being infected, and
-that plots within a row are independent of each other.
-
-\noindent Advantages of the Binomial GLMM include:
-\begin{itemize}
-\item It estimates and controls for variability between rows and blocks.
-\item It avoids confounding variety effects with row effects and correctly
-models the rows and nested within blocks.
-\item Random effects can explain extra variation beyond what the
-fixed-effect Binomial GLM can account for.
-\end{itemize}
-
-\noindent Disadvantages of this model include:
-\begin{itemize}
-\item It must be fit by numerically maximizing the likelihood with an
-iterative algorithm. Including a large number of parameters (such as
-third-order interaction coefficients) can cause the algorithm to fail
-to converge.
-\item The domain of the logit function is \((0, 1)\). If a large proportion
-of the data are zeros, meaning an infection probability is near 0, the
-algorithm will have difficulty estimating the coefficients. In this
-situation, estimates may be unreliable or the algorithm may not converge.
-\item We are not aware of any tools that can fit a Binomial GLMM with
-overdispersion. If there are additional sources of variation that the
-model does not account for, the standard error estimates will be too small.
-\end{itemize}
+that plots are independent given the other variables in the model.
 
@@
 
-\begin{figure}[b]
+\begin{figure}
 \caption{Example interaction plot using \texttt{qplot}.}
@@
 
-\noindent We noticed that infection counts of 0 were recored in 67 of the
+\noindent We noticed that infection counts of 0 were recorded in 67 of the
 180 plots in Nar's dataset. This is a large fraction of the data; if the
@@
 that treatment or variety has qualitative differences from the others
-that would cause it to have a much lower infrection rate. These data would
+that would cause it to have a much lower infection rate. These data would
 be considered as coming from a different population than the data with
@@
 
-\begin{figure}[b]
+\begin{figure}
 \caption{Example heatmap visualization for identifying patterns of zeros.}
@@
 <<startingmodel,warning=FALSE,results='hide',cache=TRUE>>=
-glm3way <- glm(cbind(infected, total) ~ block/variety + variety*n.trt*status2,
+glm3way <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
                family = quasibinomial, data = fert.sim1)
@@
 If it is decided that certain observations come from an extremely
-resistant population and can be omited from the current analysis.
+resistant population and can be omitted from the current analysis.
 
 <<novar1,warning=FALSE,results='hide',cache=TRUE>>=
-noVar1 <- glm(cbind(infected, total) ~ block/variety + variety*n.trt*status2,
+noVar1 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
               family = quasibinomial, data = fert.sim1, subset = variety!='Var1')
@@
 <<separatemodels,warning=FALSE,results='hide',cache=TRUE>>=
-onlyINOC <- glm(cbind(infected, total) ~ block/variety + variety*n.trt,
+onlyINOC <- glm(cbind(infected, total) ~ block*variety + variety*n.trt,
                 family = quasibinomial, data = fert.sim1, subset = status=='INOC')
 
-onlyCNTL <- glm(cbind(infected, total) ~ block/variety + variety*n.trt,
+onlyCNTL <- glm(cbind(infected, total) ~ block*variety + variety*n.trt,
                 family = quasibinomial, data = fert.sim1, subset = status=='CNTL')
@@
 
-<<models5,echo=FALSE,warning=FALSE,results='hide',cache=TRUE>>=
-glm3way2 <- glm(cbind(infected, total) ~ block/variety + variety*n.trt*status2,
+<<models5,echo=FALSE,warning=FALSE,message=FALSE,results='hide',cache=TRUE>>=
+glm3way2 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
                 family = binomial, data = fert2)
@@
 
-glm2way2 <- glm(cbind(infected, total) ~ block/variety + variety*n.trt + variety*status2 + n.trt*status2,
+glm2way2 <- glm(cbind(infected, total) ~ block*variety + (variety+n.trt+status2)^2,
                 family = quasibinomial, data = fert2)
@@
 Anova(glm2way2, test.statistic = 'F')
-Anova(glm2way2, glm3way2, test.statistic = 'F')
 anova(glm2way2, glm3way2)
@@
 
-<<nosnms,echo=FALSE,warning=FALSE,results='hide',cache=FALSE>>=
+<<nosnms,echo=FALSE,warning=FALSE,message=FALSE,results='hide',cache=FALSE>>=
 table(fert2$infected==0,fert2$variety)
@@
 
-<<plots3,eval=FALSE,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
-par(mfrow = c(1, 2))
+<<plots3,eval=FALSE,echo=FALSE,warning=FALSE,message=FALSE>>=
+par(mfrow = c(2, 2))
+plot(resid(glm3way2) ~ predict(glm3way2), col = variety, data = fert2)
+plot(resid(glm3way2) ~ fitted(glm3way2), col = variety, data = fert2)
 plot(resid(noSNMS3way) ~ predict(noSNMS3way), col = variety,

### 2015-10-06 leslieandkenny.Rnw
Added some GLM vs GLMM commentary

@@
 require(lattice)
-require(nlme)
 @
@@
 # Read and set up the data
-fert2 <- read.csv("fert2.csv")
+fert2 <- read.csv('fert2.csv')
 
@@
 are independent and have the same probability of being infected, and
-that plots within a row are independent of each other.
+that plots are independent given the other variables in the model.
 
-\noindent Advantages of the Binomial GLMM include:
-\begin{itemize}
-\item It estimates and controls for variability between rows and blocks.
-\item It avoids confounding variety effects with row effects and correctly
-models the rows and nested within blocks.
-\item Random effects can explain extra variation beyond what the
-fixed-effect Binomial GLM can account for.
-\end{itemize}
-
-\noindent Disadvantages of this model include:
-\begin{itemize}
-\item It must be fit by numerically maximizing the likelihood with an
-iterative algorithm. Including a large number of parameters (such as
-third-order interaction coefficients) can cause the algorithm to fail
-to converge.
-\item The domain of the logit function is \((0, 1)\). If a large proportion
-of the data are zeros, meaning an infection probability is near 0, the
-algorithm will have difficulty estimating the coefficients. In this
-situation, estimates may be unreliable or the algorithm may not converge.
-\item We are not aware of any tools that can fit a Binomial GLMM with
-overdispersion. If there are additional sources of variation that the
-model does not account for, the standard error estimates will be too small.
-\end{itemize}
+\noindent There are several reasons why we recommend using a fixed-effects
+generalized linear model instead of a generalized linear mixed model.
+We observed overdispersion in Nar's data, and there are no simple tools
+for including overdispersion in mixed models. Therefore, the standard
+errors in the mixed model output would be based on an incorrect estimate
+of the variance and might not be trustworthy. Additionally, including
+many coefficients and nested effects in a mixed model can cause
+computational problems where the software will fail to find unique
+solutions for the parameters. These problems are made worse in a Binomial
+GLMM when the counts are small, as in these data. The Binomial GLM can
+adjust for overdispersion and avoid many of the computational difficulties
+associated with the GLMM. The GLM can correctly model the nested
+design through a block by variety interaction. The only advantage of the
+GLMM is that it would estimate block- and row-level variances, but
+it is our opinion that this is outweighed by the lack of an overdispersion
+estimate and the computational difficulties.
 
@@
 that treatment or variety has qualitative differences from the others
-that would cause it to have a much lower infrection rate. \textbf{If and
-only if it is justified}, Nar could omit all of the data for this treatment
-or variety when fitting the model. The scope of inference for this model
-must then be limited to varieties and treatments that have a nonzero
-infection rate.
+that would cause it to have a much lower infection rate. These data would
+be considered as coming from a different population than the data with
+fewer zeros. \textbf{If and only if it is justified}, Nar could omit all
+of the data for this treatment or variety when fitting the model. The scope
+of inference for this model must then be limited to varieties and
+treatments that have a nonzero infection rate. A separate analysis would
+be needed for more resistant population.
 
-\begin{figure}[b]
+\begin{figure}
 \caption{Example heatmap visualization for identifying patterns of zeros.}
@@
 for spatial patterns. If correlation between nearby plots is seen,
-the model should be modyfied to account for this.
+the model should be modified to account for this.
 
@@
 
-\begin{figure}[b]
+\begin{figure}
 \caption{Example \texttt{itableplot} output.}
 <<tabplot,echo=FALSE,warning=FALSE,message=FALSE>>=
-tableplot(fert.ordered, select = c(infected, variety, block, n.trt, status),
+tableplot(fert.sim1, select = c(infected, variety, block, n.trt, status),
           nBins = 30, sortCol = variety)
@@
 
-<<models1,warning=FALSE,results='hide',cache=TRUE>>=
-glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
-                  control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
-                  family = binomial, data = fert.sim1)
+<<startingmodel,warning=FALSE,results='hide',cache=TRUE>>=
+glm3way <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
+               family = quasibinomial, data = fert.sim1)
 @
 
-We noticed that the SNMS variety had 27 out of 36 counts that were zeros.
-Code like the following
+If it is decided that certain observations come from an extremely
+resistant population and can be omitted from the current analysis:
 
-<<models2,warning=FALSE,results='hide',cache=TRUE>>=
-noVar1 <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
-                control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
-                family = binomial, data = fert.sim1, subset = variety!='Var1')
+<<novar1,warning=FALSE,results='hide',cache=TRUE>>=
+noVar1 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
+              family = quasibinomial, data = fert.sim1, subset = variety!='Var1')
 @
@@
 If there is no interest in comparing between inoculated plots and
-control plots, separate models could be fit.
+control plots, separate models could be fit:
 
-<<models3,warning=FALSE,results='hide',cache=TRUE>>=
-onlyINOC <- glmer(cbind(infected, total) ~ variety*n.trt + (1|block/row),
-                  control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
-                  family = binomial, data = fert.sim1, subset = status=='INOC')
+<<separatemodels,warning=FALSE,results='hide',cache=TRUE>>=
+onlyINOC <- glm(cbind(infected, total) ~ block*variety + variety*n.trt,
+                family = quasibinomial, data = fert.sim1, subset = status=='INOC')
 
-onlyCNTL <- glmer(cbind(infected, total) ~ variety*n.trt + (1|block/row),
-                  control = glmerControl(optimizer = 'bobyqa', optCtrl = list(maxfun = 20000)),
-                  family = binomial, data = fert.sim1, subset = status=='CNTL')
+onlyCNTL <- glm(cbind(infected, total) ~ block*variety + variety*n.trt,
+                family = quasibinomial, data = fert.sim1, subset = status=='CNTL')
 @
 
-<<models5,echo=FALSE,warning=FALSE,results='hide',cache=TRUE>>=
-glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/row),
-                  control = glmerControl(optimizer = 'bobyqa'),
-                  family = binomial, data = fert2)
-#summary(glmm3way)
+<<models5,echo=FALSE,warning=FALSE,message=FALSE,results='hide',cache=TRUE>>=
+glm3way2 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
+                family = quasibinomial, data = fert2)
+summary(glm3way2)
 
-glmm2way <- glmer(cbind(infected, total) ~ variety*n.trt + variety*status2 +
-                    n.trt*status2 + (1|block/row),
-                    control = glmerControl(optimizer = 'bobyqa'),
-                  family = binomial, data = fert2)
-#summary(glmm2way)
+glm2way2 <- glm(cbind(infected, total) ~ block*variety + (variety+n.trt+status2)^2,
+                family = quasibinomial, data = fert2)
+summary(glm2way2)
 
-#Anova(glmm3way)
-#Anova(glmm2way)
-#anova(glmm2way, glmm3way)
+Anova(glm3way2, test.statistic = 'F')
+Anova(glm2way2, test.statistic = 'F')
+anova(glm2way2, glm3way2)
 @
 
-<<plots1,eval=FALSE,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
-par(mfrow = c(1, 2))
-plot(resid(glmm3way) ~ predict(glmm3way), col = variety, data = fert.sim1)
-plot(resid(glmm3way) ~ fitted(glmm3way), col = variety, data = fert.sim1)
-legend('topright', pch = 1, col = 1:5,
-       legend = levels(fert.sim1$variety))
-@
-
-<<nosnms,echo=FALSE,warning=FALSE,results='hide',cache=FALSE>>=
+<<nosnms,echo=FALSE,warning=FALSE,message=FALSE,results='hide',cache=FALSE>>=
 table(fert2$infected==0,fert2$variety)
@@
 \end{figure}
-
-<<glmmodel1,cache=FALSE>>=
-glm1 <- glm(cbind(infected, total) ~ variety*n.trt*status2 + row,
-            family = quasibinomial, data = fert.sim1)
-glm2 <- glm(cbind(infected, total) ~ block + variety*n.trt*status2,
-            family = quasibinomial, data = fert.sim1)
-glm3 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
-            family = quasibinomial, data = fert.sim1)
-glm4 <- glm(cbind(infected, total) ~ block/row + variety*n.trt*status2,
-            family = quasibinomial, data = fert.sim1)
-glm5 <- glm(cbind(infected, total) ~ block/variety + variety*n.trt*status2,
-            family = quasibinomial, data = fert.sim1)
-
-glmnointMF <- glm(cbind(infected, total) ~ block/variety + variety*n.trt*status2,
-            family = quasibinomial, data = fert2)
-glmmnointMF <- glmer(cbind(infected, total) ~ variety + variety*n.trt*status2 + (1|block/row),
-            family = binomial, data = fert.sim1)
-@
 
@@
 explain: if no evidence of a three way interaction what can you do, how to continue selection.
+
+Adding a sentence to make sure I can push
 

### 2015-10-08 leslieandkenny.Rnw
worked on Model Refinement and Model Interpretation Sections

@@
 \usepackage{amsmath}
+\usepackage{float}
 
@@
                size = 'footnotesize')
-options(replace.assign = TRUE, width = 112, digits = 3, max.print = '72',
+options(replace.assign = TRUE, width = 112, digits = 3, max.print = '202',
         show.signif.stars = FALSE)
@@
 # Read and set up the data
-fert2 <- read.csv('fert2.csv')
+#for leslie 
+#setwd("~/Documents/ConsultingFall2015/Fall510")
+fert2 <- read.csv("fert2.csv")
 
@@
 are independent and have the same probability of being infected, and
-that plots are independent given the other variables in the model.
+that plots are independent given the other variables in the model. \\
 
@@
 may be present, and for identifying patterns that may represent violations
-of assumptions or potential computational problems.
+of assumptions or potential computational problems. \\
 
@@
 it may be impossible to estimate coefficients for factor levels where
-infection rates are nearly zero.
+infection rates are nearly zero. \\
 
@@
 The most serious problems will come from the zeros, so we chose a contrasting
-color (black) for plots with infection counts of 0.
+color (black) for plots with infection counts of 0.\\
 
@@
 the the three-way interaction involving fall nitrogen application and
-variety 1.
+variety 1. \\
 
@@
 \subsection{Model Fitting and Possible Issues}
+\label{model}
 
@@
 <<startingmodel,warning=FALSE,results='hide',cache=TRUE>>=
-glm3way <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
+#this forces the constraint that the block effects sum to 0
+contrasts(fert.sim1$block) <- contr.sum(6, contrast=TRUE)
+glm3way <- glm(cbind(infected, total) ~ block/variety + variety*n.trt*status2,
                family = quasibinomial, data = fert.sim1)
@@
 
-\begin{figure}
-\caption{The fitted model can be nicely summarized with plots from the \texttt{effects} package.}
-<<plots2,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5>>=
+\subsection{Model Refinement}
+
+In this section, we describe how to go through the process of model selection.
+As discussed in the previous section, we recommend starting with the most complicated $3$ way interaction model and then assessing which terms can be removed.
+In general, start by considering removal of the highest order terms.
+\\
+
+\noindent We'll start by assessing evidence for the three-way interaction term.
+One way to consider it's removal would be to fit a model without the three-way interaction, and then conduct a drop-in-deviance F-test comparing the full and reduced models, as you learned to do in Stat $512$.
+This is method can be time consuming, so instead we recommend using the \verb+Anova+ function in the \verb+car+ package, and choosing the option \verb+test="F"+.
+This provides a {\bf Type II} sums of squares table, and the tests provided are similar to extra sums of squares F-tests based on the Pearson residuals (provide reference here to info about pearson resids vs deviance resids).
+This terms in this table are tested conditional upon terms of the same and lower order being in the model.
+For example, the \verb+n.trt:status2+ row in the table tests for removal of the $n.trt*status2$ two-way interaction, conditional upon all other two way interactions and main effects being in the model.
+As a result, this table can be used to assess evidence for removal of each term.
+
+<<refine, echo=TRUE>>=
+Anova(glm3way, test="F")
+@
+
+\noindent Here is how we would interpret the results in each row of the table.
+Note, it is not appropriate to consider removal of any of the terms involving block.
+Block to block variability is expected because it was included as a design variable, and so it should be included in the model.
+The $block*variety$ interaction also needs to be included to account for the nesting structure in the design.
+\begin{enumerate}
+\item \verb+variety:n.trt:status2+ ``There is no evidence that the interaction between variety and nitrogen application time depends on inoculation status (p-value$=0.388$ from F-stat$=1.07$ on $8$ and $125$ df).''
+\item \verb+n.trt:status2+ ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection depends on inoculation status after controlling for all other two way interactions in the model (p-value$=0.368$ from F-stat$=1.01$ on $2$ and $133$ df).
+\item \verb+variety:status2+ ``There is no evidence that the relationship between variety and the odds of virus infection depends on inoculation status after controlling for all other two way interactions in the model (p-value$=0.288$ from F-stat$=1.26$ on $4$ and $133$ df).
+\item \verb+variety:n.trt+ ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection depends on variety after controlling for all other two way interactions in the model (p-value$=0.453$ from F-stat$=0.98$ on $8$ and $133$ df).
+\end{enumerate}
+
+\noindent If no evidence is found for any of the interactions, the simplest model we recommend fitting should contain the $block*variety$ interaction and all main effects (shown below).
+If this model is selected, each effect is then interpreted after controlling for the effects of the other variables in the model.
+For example, to assess evidence for the \verb+n.trt+ effect in this model, we would say, ``There is no evidence that the odds of virus infection depends on nitrogen application time after controlling for block, variety, inoculation status, and the
+$block*variety$ interaction.''
+
+<<glmnointeractions, echo=TRUE>>=
+glmnointeractions <- glm(cbind(infected, total) ~ block/variety + variety+n.trt 
+                         + status2,
+               family = quasibinomial, data = fert.sim1)
+Anova(glmnointeractions, test="F")
+@
+
+\noindent The next section describes how to interpret effects once a model is selected.
+
+<<dropindevftest, echo=FALSE>>=
+glm.no3way <- glm(cbind(infected, total) ~ block/variety + variety*n.trt + n.trt*status2 + variety*status2,
+               family = quasibinomial, data = fert.sim1)
+#drop in dev F-test
+#((196-185)/8)/(185/125) 
+#0.929
+#1-pf(0.929, 11, 125)
+#0.512
+@
+
+\subsection{Interpretation}
+
+\noindent In this section, we describe how to interpret effects if evidence of a three-way interaction is found.
+The effects plots in the \verb+effects+ packages provide nice visuals for displaying estimated probabilities of infection across levels of mulitiple variables.
+
+\begin{figure}[!h]
+<<plots2,echo=FALSE,warning=FALSE,message=FALSE,out.height='0.5\\linewidth',fig.height=5, fig.align='center'>>=
 require(effects)
@@
 @
+\caption{The fitted model can be nicely summarized with plots from the \texttt{effects} package.}
 \end{figure}
 
-\subsection{Model Refinement}
+\noindent The \verb+Effect+ function is convenient because it provides the estimated probabilities at each treatment combination, averaged over all blocks.
+The results are  provided in a organized table (see below).
 
-Recommend to use Chi Squared tests or F-tests? I think F-tests but not exactly sure what the F-test is giving or where its coming from.
-
-Correct df? different from lme and glmer...
-
-<<nosnms3way, echo=FALSE>>=
-Anova(noSNMS3way)
-anova(noSNMS3way)
-#1-pf(2.29, 6, 118)
-
-AIC(noSNMS3way, noSNMS2way)
-anova(noSNMS3way, noSNMS2way)
-
-#lme model to get correct df for f tests
-#why is den df now 100?
-require(nlme, quietly=T)
-noSNMS3waylme <- lme(infected ~ variety*n.trt*status,
-                    random=~1|block/row, method="ML",
-                     data = subset(fert2, variety != "SNMS"))
-
-glm3way <- glm(cbind(infected, total) ~ variety*n.trt*status+row, family="binomial", data=subset(fert2, variety != "SNMS"))
+<<summary, echo=TRUE>>=
+Effect(c('variety','n.trt','status2'), glm3way, se=TRUE)
 @
 
-explain: if no evidence of a three way interaction what can you do, how to continue selection.
+\noindent We recommend displaying the effects plot and output and then describing the effects of interest in words.
+When describing the effects in words, we recommend describing the effects in terms of multiplicative changes in the odds of virus infection.
+It is more straightforward to calculate confidence intervals when effects are described are this scale rather than the probability scale.
+We recommend changing the reference level of the model to find the estimates and confidence intervals for the effects of interest.
+Notice that when we fit the model in Section \ref{model}, we specified the constraint that the sum of the block effects would add to $0$.
+As a result, the estimated effects given in the model summary are the effects averaged over all blocks.
+\\
 
-Adding a sentence to make sure I can push
+\noindent You can start by looking at the summary of the model, \verb+summary(glm3way)+.
+The coefficients of interest will be those found in the rows \verb+n.trtEARLY SPRING+ and \verb+n.trtLATE SPRING+.
+Below we show how to interpret these coefficients in words, and how the meaning changes when the reference level is manipulated.
 
-\subsection{Interpretation}
+<<first, echo=TRUE, message=FALSE>>=
+#use the default reference level to describe the effects across 
+#nitrogen application for inoculated plants of variety 1 (averaged over blocks)
+exp(coef(glm3way)[11])
+exp(confint(glm3way, 11))
+exp(coef(glm3way)[12])
+exp(confint(glm3way, 12))
+@
 
-Explain how to interpret with three way interaction (essentially interpret two way interaction for each inoculated and control)
+\noindent For these data, we would describe the nitrogen application effect for inoculated plants of variety $1$ as follows:
+\begin{quote}
+For inoculated plants of variety $1$, the odds of virus infection when nitrogen is applied in the early spring is estimated to be $3.01$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from  $0.295$ to $104.9$ times.
+For inoculated plants of variety $1$, the odds of virus infection when nitrogen is applied in the late spring is estimated to be $5.05$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from $0.652$ to $166.4$ times.
+\end{quote}
+
+<<second, echo=TRUE, message=FALSE, eval=FALSE>>=
+#suppose we want to the effects of introgen application for 
+#inoculated plants of variety 3
+#first relevel
+var3 <- relevel(fert.sim1$variety, "Var3")
+#then fit model with var3
+glm.refvar3 <- glm(cbind(infected, total) ~ block/var3 + var3*n.trt*status2,
+               family = quasibinomial, data = fert.sim1)
+#interpret results
+exp(coef(glm.refvar3)[11])
+exp(confint(glm.refvar3, 11))
+exp(coef(glm.refvar3)[12])
+exp(confint(glm.refvar3, 12))
+
+#now suppose we want to the effects of introgen application for 
+#non-inoculated plants of variety 3
+#first relevel the status variable
+statuscntl <- relevel(fert.sim1$status2, "CNTL")
+#then fit model with var3
+glm.refvar3cntl <- glm(cbind(infected, total) ~ block/var3 + var3*n.trt*statuscntl,
+               family = quasibinomial, data = fert.sim1)
+#interpret results
+exp(coef(glm.refvar3cntl)[11])
+exp(confint(glm.refvar3cntl, 11))
+exp(coef(glm.refvar3cntl)[12])
+exp(confint(glm.refvar3cntl, 12))
+@
+
+\noindent We would describe the nitrogen application effect for inoculated plants of variety $3$ as follows:
+\begin{quote}
+For inoculated plants of variety $3$, the odds of virus infection when nitrogen is applied in the early spring is estimated to be about equal to the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from  $0.191$ to $5.28$ times.
+For inoculated plants of variety $3$, the odds of virus infection when nitrogen is applied in the late spring is estimated to be $1.24$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from $0.269$ to $6.201$ times.
+\end{quote}
+
+\noindent We would describe the nitrogen application effect for non-inoculated plants of variety $3$ as follows:
+\begin{quote}
+For non-inoculated plants of variety $3$, the odds of virus infection when nitrogen is applied in the early spring is estimated to be $2.3$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from  $0.797$ to $7.72$ times.
+For inoculated plants of variety $3$, the odds of virus infection when nitrogen is applied in the late spring is estimated to be $1.45$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from $0.446$ to $5.16$ times.
+\end{quote}
 

### 2015-10-09 leslieandkenny.Rnw
minor changes before submission

@@
 He is currently working on several projects regarding the spread and development of the Wheat Streak Mosaic Virus (WSMV) in Winter Wheat.
-Nar has taken Statistics $511$ and $512$ here at Montana State University, and he is currently sitting in Mixed Models.
+Nar has taken Statistics $511$ and $512$ here at Montana State University, and he is currently sitting in on Mixed Models.
 Nar has asked for our help in building a model, fitting it in R, and interpreting the results.
@@
 Nar conducted this experiment in a MSU field at the base of the Bridger Mountains.
-He chose one area in the field and divided it into six $31.5$ by $27.5$ meter blocks.
+He divided the field into six $31.5$ by $27.5$ meter blocks.
 The blocks were then divided into five rows, with $5$ meters of space between each row.
@@
 
-\noindent Plots that were assigned to the inoculated treatment had five infected plants transplanted to the middle of the plot.
+\noindent Plots assigned to the inoculated treatment had five infected plants transplanted to the middle of the plot.
 These plants were infected in the greenhouse by clipping an infected leaf to the healthy plant.
-The plants are infected in the tillering stage of their life cycle at an age of about one month.
+The plants were infected in the tillering stage of their life cycle at an age of about one month.
 The Wheat Streak Mosaic Virus is transmitted by the wheat curl mite, tiny organisms that are nearly invisible to the naked eye.
@@
 It is conceivable that the mites on the infected plants introduced to the inoculated plots could move to healthy plants within that plot.
-But, the researchers say that it is nearly impossible for the mites to move from an inoculated plot to a non-inoculated plot with such a low population of mites.
+But, Nar says that it is nearly impossible for the mites to move from an inoculated plot to a non-inoculated plot with such a low population of mites.
 Plants with the virus in non-inoculated plots are assumed to have been infected by mites in the surrounding environment.
@@
 
-\noindent All plots were planted in mid September and allowed to grow over the winter.
+\noindent All plots were planted in mid-September and allowed to grow over the winter.
 The following spring and summer, $30$ leaves were picked from different plants within each plot and taken to the lab to be tested for presence of the virus.
@@
 combination of factor levels. The \verb|qplot| function in the
-\verb|ggplot2| package is a simple way to do this.
+\verb|ggplot2| package is a simple way to do this. \\
 
@@
 treatments that have a nonzero infection rate. A separate analysis would
-be needed for more resistant population.
+be needed for more resistant population. \\
 
@@
 for spatial patterns. If correlation between nearby plots is seen,
-the model should be modified to account for this.
+the model should be modified to account for this. \\
 
@@
 
-The following code will fit the three-way interaction model:
+\noindent The following code will fit the three-way interaction model:
 
@@
 
-If it is decided that certain observations come from an extremely
+\noindent If it is decided that certain observations come from an extremely
 resistant population and can be omitted from the current analysis:
@@
 
-If there is no interest in comparing between inoculated plots and
+\noindent If there is no interest in comparing between inoculated plots and
 control plots, separate models could be fit:
@@
 In this section, we describe how to go through the process of model selection.
-As discussed in the previous section, we recommend starting with the most complicated $3$ way interaction model and then assessing which terms can be removed.
+As discussed in the previous section, we recommend starting with the most complicated three-way interaction model and then assessing which terms can be removed.
 In general, start by considering removal of the highest order terms.
@@
 One way to consider it's removal would be to fit a model without the three-way interaction, and then conduct a drop-in-deviance F-test comparing the full and reduced models, as you learned to do in Stat $512$.
-This is method can be time consuming, so instead we recommend using the \verb+Anova+ function in the \verb+car+ package, and choosing the option \verb+test="F"+.
+This method can be time consuming, so instead we recommend using the \verb+Anova+ function in the \verb+car+ package with the option \verb+test="F"+.
 This provides a {\bf Type II} sums of squares table, and the tests provided are similar to extra sums of squares F-tests based on the Pearson residuals (provide reference here to info about pearson resids vs deviance resids).
-This terms in this table are tested conditional upon terms of the same and lower order being in the model.
+The tests in this table are conditional upon terms of the same and lower order being in the model.
 For example, the \verb+n.trt:status2+ row in the table tests for removal of the $n.trt*status2$ two-way interaction, conditional upon all other two way interactions and main effects being in the model.
@@
 \begin{enumerate}
-\item \verb+variety:n.trt:status2+ ``There is no evidence that the interaction between variety and nitrogen application time depends on inoculation status (p-value$=0.388$ from F-stat$=1.07$ on $8$ and $125$ df).''
-\item \verb+n.trt:status2+ ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection depends on inoculation status after controlling for all other two way interactions in the model (p-value$=0.368$ from F-stat$=1.01$ on $2$ and $133$ df).
-\item \verb+variety:status2+ ``There is no evidence that the relationship between variety and the odds of virus infection depends on inoculation status after controlling for all other two way interactions in the model (p-value$=0.288$ from F-stat$=1.26$ on $4$ and $133$ df).
+\item \verb+variety:n.trt:status2+ ``There is no evidence that the interaction between variety and nitrogen application time differs between inoculated and control plots (p-value$=0.388$ from F-stat$=1.07$ on $8$ and $125$ df).''
+\item \verb+n.trt:status2+ ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection differs between inoculated and control plots after controlling for all other two way interactions in the model (p-value$=0.368$ from F-stat$=1.01$ on $2$ and $133$ df).
+\item \verb+variety:status2+ ``There is no evidence that the relationship between variety and the odds of virus infection differs between inoculated and control plots after controlling for all other two way interactions in the model (p-value$=0.288$ from F-stat$=1.26$ on $4$ and $133$ df).
 \item \verb+variety:n.trt+ ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection depends on variety after controlling for all other two way interactions in the model (p-value$=0.453$ from F-stat$=0.98$ on $8$ and $133$ df).
@@
 For non-inoculated plants of variety $3$, the odds of virus infection when nitrogen is applied in the early spring is estimated to be $2.3$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from  $0.797$ to $7.72$ times.
-For inoculated plants of variety $3$, the odds of virus infection when nitrogen is applied in the late spring is estimated to be $1.45$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from $0.446$ to $5.16$ times.
+For non-inoculated plants of variety $3$, the odds of virus infection when nitrogen is applied in the late spring is estimated to be $1.45$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from $0.446$ to $5.16$ times.
 \end{quote}

### 2015-10-09 leslieandkenny.Rnw
Not sure what changed but git is telling me to commit
First draft ready to turn in

@@
 fert.sim1$status2 <- relevel(fert.sim1$status, 'INOC')
+
+# This forces the constraint that the block effects sum to 0
+contrasts(fert.sim1$block) <- contr.sum(6, contrast=TRUE)
 @
@@
 In the first year of the study, $2013-2014$, all plots were sampled on both dates.
-In the second year of the study, only plots with susceptible varieties were sampled on the first date, and all plots were sampled on the second date The advice given in this report is only for analysis of the data collected on the second date in July.
+In the second year of the study, only plots with susceptible varieties were sampled on the first date, and all plots were sampled on the second date.
+The advice given in this report is only for analysis of the data collected on the second date in July.
 After the leaves were picked in the field, they were sent to the lab and screened for the virus via the ELISA procedure.
@@
 interaction to control for differences between blocks and between rows
-in the same block.
+in the same block.\\
 
@@
 \noindent There are several reasons why we recommend using a fixed-effects
-generalized linear model instead of a generalized linear mixed model.
-We observed overdispersion in Nar's data, and there are no simple tools
+generalized linear model instead of a generalized linear mixed model;
+these reasons are related to overdispersion and computational issues. \\
+
+\noindent Since the inoculated plants were planted in the center of the
+plots, we expect to see spatial dependence where plants near the centers
+are more likely to become infected than plants further from the centers.
+Then individual plants could have infection probabilities that differ from
+the infection rate estimated for the plot they are in. This would result
+in the data could having more variation than what the Binomial model
+accounts for, a situation known as overdispersion. We observed
+overdispersion in Nar's data, and there are no simple tools
 for including overdispersion in mixed models. Therefore, the standard
 errors in the mixed model output would be based on an incorrect estimate
-of the variance and might not be trustworthy. Additionally, including
-many coefficients and nested effects in a mixed model can cause
-computational problems where the software will fail to find unique
-solutions for the parameters. These problems are made worse in a Binomial
-GLMM when the counts are small, as in these data. The Binomial GLM can
-adjust for overdispersion and avoid many of the computational difficulties
-associated with the GLMM. The GLM can correctly model the nested
-design through a block by variety interaction. The only advantage of the
-GLMM is that it would estimate block- and row-level variances, but
-it is our opinion that this is outweighed by the lack of an overdispersion
-estimate and the computational difficulties.
+of the variance and might not be trustworthy. This problem is easily
+remedied by using a Quasibinomial fixed-effects GLM model. \\
+
+\noindent Additionally, including many coefficients and nested effects
+in a mixed model can cause computational problems where the software will
+fail to find unique solutions for the parameters. These problems are made
+worse in a Binomial GLMM when the counts are small, as in these data. The
+Binomial GLM can avoid many of the computational difficulties associated
+with the GLMM. The GLM can correctly model the nested design through a
+block by variety interaction. \\
+
+\noindent The only advantage of the GLMM is that it would estimate block-
+and row-level variances, but it is our opinion that this is outweighed by
+the lack of an overdispersion estimate and the computational difficulties.
 
@@
 
-\begin{figure}
+\begin{figure}[h!]
 \caption{Example interaction plot using \texttt{qplot}.}
-<<interactionplot,echo=FALSE,message=FALSE>>=
+<<interactionplot,echo=FALSE,message=FALSE,out.height='0.3\\linewidth',fig.height=3>>=
 require(ggplot2)
-qplot(x = n.trt, y = infected, geom = 'boxplot', color = status,
+qplot(x = n.trt, y = infected, geom = c('boxplot', 'point'), color = status,
       facets = .~variety, data = fert.sim1) +
@@
 treatments that have a nonzero infection rate. A separate analysis would
-be needed for more resistant population. \\
+be needed for the more resistant population. \\
 
-\begin{figure}
+\begin{figure}[h!]
 \caption{Example heatmap visualization for identifying patterns of zeros.}
-<<heatmap,echo=FALSE,message=FALSE>>=
+<<heatmap,echo=FALSE,message=FALSE,out.height='0.3\\linewidth',fig.height=3>>=
 ## Heatmap with meaningful ordering
@@
 # Plot the heatmap, with zeros in black and segments separating the blocks
-par(mar = c(5, 10, 6, 2)) # Set big margins
+par(mar = c(3, 10, 6, 2)) # Set big margins
 image(z = infected.arranged, y = 1:6,
@@
 # Legend
-par(mar = c(5, 1, 6, 2))
+par(mar = c(3, 1, 6, 2))
 image(y = seq(-0.5, 30.5, 1), z = matrix(0:30, nrow = 1), axes = FALSE, ylab = '',
@@
 
-\begin{figure}
+\begin{figure}[h!]
 \caption{Example \texttt{itableplot} output.}
@@
 
-\subsection{Model Fitting and Possible Issues}
+\subsection{Model Fitting}
 \label{model}
@@
 <<startingmodel,warning=FALSE,results='hide',cache=TRUE>>=
-#this forces the constraint that the block effects sum to 0
-contrasts(fert.sim1$block) <- contr.sum(6, contrast=TRUE)
 glm3way <- glm(cbind(infected, total) ~ block/variety + variety*n.trt*status2,
@@
 This method can be time consuming, so instead we recommend using the \verb+Anova+ function in the \verb+car+ package with the option \verb+test="F"+.
-This provides a {\bf Type II} sums of squares table, and the tests provided are similar to extra sums of squares F-tests based on the Pearson residuals (provide reference here to info about pearson resids vs deviance resids).
+This provides a {\bf Type II} sums of squares table, and the tests provided are similar to extra sums of squares F-tests based on the Pearson residuals.
+%(provide reference here to info about pearson resids vs deviance resids)
 The tests in this table are conditional upon terms of the same and lower order being in the model.
@@
 \noindent In this section, we describe how to interpret effects if evidence of a three-way interaction is found.
-The effects plots in the \verb+effects+ packages provide nice visuals for displaying estimated probabilities of infection across levels of mulitiple variables.
+The effects plots in the \verb+effects+ packages provide nice visuals for displaying estimated probabilities of infection across levels of multiple variables.
 
@@
 result, they present technical and computational problems for analysis.
-In similar future studies, random sampling should be used within each
-plot and the sample size should be large enough that the researchers
-expect at least one infected leaf to be found in each plot.
+In similar future studies, the sample size should be large enough that
+the researchers expect at least one infected leaf to be found in each plot.
+Additionally we recommend placing the inoculated plants at random locations
+within the plots, and collecting leaves from a simple random sample of
+locations in each plot.
 
@@
 \noindent Sloderbeck, J., Michaud, P., Whitworth, Robert. "Wheat Pests." CurlMite. Kansas State University, 1 May 2008. Web. 18 Sept. 2015.\\
-\verb|http://entomology.k-state.edu/extension/insect-information/crop-pests/wheat/curlmite.html|.
+{\footnotesize\verb|http://entomology.k-state.edu/extension/insect-information/crop-pests/wheat/curlmite.html|.}
 

### 2015-10-19 leslieandkenny.Rnw
Started adding QuasiBinom explanation, made code style more consistent in places

@@
         show.signif.stars = FALSE)
+knit_theme$set('print')
 
@@
 # Correctly order nitrogen application levels
-fert2$n.trt <- relevel(fert2$n.application, 'FALL')
+fert2$n.trt <- relevel(fert2$n.application, "FALL")
 
 # If status=INOC is the most interesting case, make INOC the baseline
-fert2$status2 <- relevel(fert2$status, 'INOC')
+fert2$status2 <- relevel(fert2$status, "INOC")
 
 # If YSTN is the variety we want to know about the most, make it baseline
-fert2$variety2 <- relevel(fert2$variety, 'YSTN')
+fert2$variety2 <- relevel(fert2$variety, "YSTN")
 
@@
 set.seed(983724)
-fert.sim1 <- data.frame(
-  'block' = c(rep('A', 30), rep('B', 30), rep('C', 30), rep('D', 30), rep('E', 30), rep('F', 30)),
-  'variety' = c(rep('Var1', 6), rep('Var2', 6), rep('Var3', 6), rep('Var4', 6), rep('Var5', 6)),
-  'n.application' = rep(c('FALL', 'FALL', 'EARLY SPRING', 'EARLY SPRING', 'LATE SPRING', 'LATE SPRING'), 30),
-  'status' = rep(c('CNTL', 'INOC'), 90),
-  'total' = rep(30, 180),
-  'infected' = rnbinom(180, 4, 7/9))
-#  'infected' = sample(fert2$infected, 180, replace = TRUE))
+fert.sim1 <- data.frame(expand.grid(
+  "status" = c("CNTL", "INOC"),
+  "n.application" = c("FALL", "EARLY SPRING", "LATE SPRING"),
+  "variety" = c("Var1", "Var2", "Var3", "Var4", "Var5"),
+  "block" = LETTERS[1:6],
+  "total" = 30),
+  "infected" = rnbinom(180, 4, 7/9))
+#  "infected" = sample(fert2$infected, 180, replace = TRUE))
 @
@@
 # Correctly order nitrogen application levels, the default is alphabetical
-fert.sim1$n.trt <- relevel(fert.sim1$n.application, 'FALL')
+fert.sim1$n.trt <- relevel(fert.sim1$n.application, "FALL")
 
 # If status=INOC is the most interesting case, make INOC the baseline
-fert.sim1$status2 <- relevel(fert.sim1$status, 'INOC')
+fert.sim1$status2 <- relevel(fert.sim1$status, "INOC")
 
@@
 $$
-y_i\sim\mathrm{Binomial}(30, p_i),
+y_i\sim\mathrm{overdispersed Binomial}(30, p_i, \omega),
 $$
@@
 variety, nitrogen application timing, and inoculation status of
-observation \(i\).
+observation \(i\),
+\item \(\omega\) is an overdispersion parameter, so that the standard
+deviation of the response in plot \(i\) is \(\sqrt{30\omega p_i(1-p_i)}\).
 \end{itemize}
@@
 require(ggplot2)
-qplot(x = n.trt, y = infected, geom = c('boxplot', 'point'), color = status,
+qplot(x = n.trt, y = infected, geom = c("boxplot", "point"), color = status,
       facets = .~variety, data = fert.sim1) +
@@
       # Use black for 0, and use heatmap colors (red-orange-yellow-white) for 1 to 30
-      col = c('black', heat.colors(30)), zlim = c(0, 30),
+      col = c("black", heat.colors(30)), zlim = c(0, 30),
 
       # Don't automatically create axes or labels
-      xlab = '', ylab = '', yaxt = 'n', xaxt = 'n')
+      xlab = "", ylab = "", yaxt = "n", xaxt = "n")
 
 # Use white line segments to visually separate the varieties
-segments(x0 = 1.5:4.5, y0 = 0.5, y1 = 6.5, col = 'white')
+segments(x0 = 1.5:4.5, y0 = 0.5, y1 = 6.5, col = "white")
 
 # Place a title at the top, and label Varieties, treatment:status levels, and blocks around the image
-title('Infection Counts', line = 4)
+title("Infection Counts", line = 4)
 axis(3, labels = rep(levels(fert.ordered$block), 5), cex.axis = 0.75,
@@
 par(mar = c(3, 1, 6, 2))
-image(y = seq(-0.5, 30.5, 1), z = matrix(0:30, nrow = 1), axes = FALSE, ylab = '',
-      col = c('#000000', heat.colors(30)), zlim = c(0, 30))
-title('Legend', line = 1.5)
+image(y = seq(-0.5, 30.5, 1), z = matrix(0:30, nrow = 1), axes = FALSE, ylab = "",
+      col = c("#000000", heat.colors(30)), zlim = c(0, 30))
+title("Legend", line = 1.5)
 axis(4)
@@
 tableplot(fert.sim1, select = c(infected, variety, block, n.trt, status),
-          nBins = 30, sortCol = variety)
+          nBins = 30, sortCol = "variety")
 @
@@
 noVar1 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
-              family = quasibinomial, data = fert.sim1, subset = variety!='Var1')
+              family = quasibinomial, data = fert.sim1, subset = variety!="Var1")
 @
@@
 onlyINOC <- glm(cbind(infected, total) ~ block*variety + variety*n.trt,
-                family = quasibinomial, data = fert.sim1, subset = status=='INOC')
+                family = quasibinomial, data = fert.sim1, subset = status=="INOC")
 
 onlyCNTL <- glm(cbind(infected, total) ~ block*variety + variety*n.trt,
-                family = quasibinomial, data = fert.sim1, subset = status=='CNTL')
+                family = quasibinomial, data = fert.sim1, subset = status=="CNTL")
 @
@@
 
-Anova(glm3way2, test.statistic = 'F')
-Anova(glm2way2, test.statistic = 'F')
+Anova(glm3way2, test.statistic = "F")
+Anova(glm2way2, test.statistic = "F")
 anova(glm2way2, glm3way2)
@@
 noSNMS3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
-                    (1|block/row), subset = variety != 'SNMS',
-                    control = glmerControl(optimizer = 'bobyqa'),
+                    (1|block/row), subset = variety != "SNMS",
+                    control = glmerControl(optimizer = "bobyqa"),
                     family = binomial, data = fert2)
@@
 noSNMS2way <- glmer(cbind(infected, total) ~ variety*n.trt + variety*status2 +
-                    n.trt*status2 + (1|block/row), subset = variety != 'SNMS',
-                    control = glmerControl(optimizer = 'bobyqa'),
+                    n.trt*status2 + (1|block/row), subset = variety != "SNMS",
+                    control = glmerControl(optimizer = "bobyqa"),
                     family = binomial, data = fert2)
@@
 plot(resid(noSNMS3way) ~ predict(noSNMS3way), col = variety,
-     data = fert2, subset = variety!='SNMS')
+     data = fert2, subset = variety!="SNMS")
 plot(resid(noSNMS3way) ~ fitted(noSNMS3way), col = variety,
-     data = fert2, subset = variety!='SNMS')
-legend('topright', pch = 1, col = 1:5,
+     data = fert2, subset = variety!="SNMS")
+legend("topright", pch = 1, col = 1:5,
        legend = levels(fert2$variety))
@@
 <<glmnointeractions, echo=TRUE>>=
-glmnointeractions <- glm(cbind(infected, total) ~ block/variety + variety+n.trt
-                         + status2,
+glmnointeractions <- glm(cbind(infected, total) ~ block*variety + variety + n.trt + status2,
                family = quasibinomial, data = fert.sim1)
@@
 <<dropindevftest, echo=FALSE>>=
-glm.no3way <- glm(cbind(infected, total) ~ block/variety + variety*n.trt + n.trt*status2 + variety*status2,
+glm.no3way <- glm(cbind(infected, total) ~ block*variety + variety*n.trt + n.trt*status2 + variety*status2,
                family = quasibinomial, data = fert.sim1)
@@
 require(effects)
-plot(Effect(c('variety','n.trt','status2'), glm3way),
-     multiline = TRUE, type = 'response', se = TRUE, ci.style='bars',
-     x.var = 'n.trt', rotx = 45, layout = c(5, 1),
-     main = 'Variety/Treatment/Inoculation Effects Plot',
-     xlab = 'Nitrogen Application Date', ylab = 'Infection Probability',
-     key.args = list(title = 'Inoculation Status'))
+plot(Effect(c("variety", "n.trt", "status2"), glm3way),
+     multiline = TRUE, type = "response", se = TRUE, ci.style="bars",
+     x.var = "n.trt", rotx = 45, layout = c(5, 1),
+     main = "Variety/Treatment/Inoculation Effects Plot",
+     xlab = "Nitrogen Application Date", ylab = "Infection Probability",
+     key.args = list(title = "Inoculation Status"))
 @
@@
 <<summary, echo=TRUE>>=
-Effect(c('variety','n.trt','status2'), glm3way, se=TRUE)
+Effect(c("variety", "n.trt", "status2"), glm3way, se=TRUE)
 @

### 2015-10-19 leslieandkenny.Rnw
Tried some more glmms

@@
 \author{Leslie Gains-Germain, Kenny Flagg}
-\date{Spring 2015}
+\date{Fall 2015}
 
@@
 
+# This forces the constraint that the block effects sum to 0
+contrasts(fert2$block) <- contr.sum(6, contrast=TRUE)
+
 # Correctly order nitrogen application levels
@@
 
+# Create a plot factor
+fert2$plot <- factor(1:180)
+
 # Combine nitrogen and inoc/cntl (for the heatmap labels)
-#fert2$n.status <- with(fert2, interaction(n.trt, status))
+fert2$n.status <- with(fert2, interaction(n.trt, status))
 @
@@
 <<tabplot,echo=FALSE,warning=FALSE,message=FALSE>>=
+#Warning message:
+#In tableplot_checkCols(substitute(sortCol), colNames) :
+#  Sorting on multiple columns not supported anymore. Only first column is taken.
 tableplot(fert.sim1, select = c(infected, variety, block, n.trt, status),
@@
 anova(glm2way2, glm3way2)
+@
+
+<<glmms,echo=FALSE,warning=FALSE,message=FALSE,results='hide',cache=TRUE>>=
+glmBin <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
+              family = binomial, data = fert2)
+# Warning message:
+# glm.fit: fitted probabilities numerically 0 or 1 occurred
+# Residual deviance: 427.939  on 125  degrees of freedom
+# 427.939 / 125 = 3.423512
+
+glmQBin <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
+               family = quasibinomial, data = fert2)
+# No warnings
+# Residual deviance: 427.939  on 125  degrees of freedom
+# (Dispersion parameter for quasibinomial family taken to be 3.2817219)
+
+glmmVar <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
+                   (1|block/variety),
+                 control = glmerControl(optimizer = "bobyqa",
+                                        optCtrl = list("maxfun" = 20000)),
+                 family = binomial, data = fert2)
+# Warning messages:
+# 1: In checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv,  :
+#   unable to evaluate scaled gradient
+# 2: In checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv,  :
+#   Model failed to converge: degenerate  Hessian with 2 negative eigenvalues
+# variety:block (Intercept) 0.46633  0.68289
+# block         (Intercept) 0.00000  0.00000
+# varietySNMS:n.trtLATE SPRING               -18.53042 2535.76041
+# varietySNMS:status2CNTL                    -18.50610 2527.55827
+# varietySNMS:n.trtEARLY SPRING:status2CNTL   19.11619 2527.55834
+# varietySNMS:n.trtLATE SPRING:status2CNTL    19.07429 4381.94303
+# deviance df.resid
+#    865.7      148
+# 865.7 / 148 = 5.8493243
+
+glmmVarPlot <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
+                       (1|block/variety) + (1|plot),
+                     control = glmerControl(optimizer = "bobyqa",
+                                            optCtrl = list("maxfun" = 20000)),
+                     family = binomial, data = fert2)
+# Warning messages:
+# 1: In vcov.merMod(object, use.hessian = use.hessian) :
+#   variance-covariance matrix computed from finite-difference Hessian is
+# not positive definite or contains NA values: falling back to var-cov estimated from RX
+# 2: In vcov.merMod(object, correlation = correlation, sigm = sig) :
+#   variance-covariance matrix computed from finite-difference Hessian is
+# not positive definite or contains NA values: falling back to var-cov estimated from RX
+# plot          (Intercept) 1.03818  1.0189
+# variety:block (Intercept) 0.28324  0.5322
+# block         (Intercept) 0.00000  0.0000
+# varietySNMS:n.trtLATE SPRING                -20.10372  8008.61084
+# varietySNMS:status2CNTL                     -19.96154  7604.01229
+# varietySNMS:n.trtEARLY SPRING:status2CNTL    20.93076  7604.01238
+# varietySNMS:n.trtLATE SPRING:status2CNTL     20.61822 13745.14025
+# deviance df.resid
+#    726.8      147
+# 726.8 / 147 = 4.9442177
+
+glmmRow <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
+                   (1|block/row),
+                 control = glmerControl(optimizer = "bobyqa",
+                                        optCtrl = list("maxfun" = 20000)),
+                 family = binomial, data = fert2)
+# Warning messages:
+# 1: In checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv,  :
+#   unable to evaluate scaled gradient
+# 2: In checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv,  :
+#   Model failed to converge: degenerate  Hessian with 3 negative eigenvalues
+# row:block (Intercept) 0.46633  0.68289
+# block     (Intercept) 0.00000  0.00000
+# varietySNMS:n.trtLATE SPRING               -18.52017 2522.79684
+# varietySNMS:status2CNTL                    -18.50714 2528.86662
+# varietySNMS:n.trtEARLY SPRING:status2CNTL   19.11723 2528.86668
+# varietySNMS:n.trtLATE SPRING:status2CNTL    19.06600 4374.54075
+# deviance df.resid
+#    865.7      148
+# 865.7 / 148 = 5.8493243
+
+glmmRowPlot <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
+                       (1|block/row) + (1|plot),
+                     control = glmerControl(optimizer = "bobyqa",
+                                            optCtrl = list("maxfun" = 20000)),
+                     family = binomial, data = fert2)
+# Warning messages:
+# 1: In checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv,  :
+#   unable to evaluate scaled gradient
+# 2: In checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv,  :
+#   Model failed to converge: degenerate  Hessian with 2 negative eigenvalues
+# plot      (Intercept) 1.03818  1.0189
+# row:block (Intercept) 0.28324  0.5322
+# block     (Intercept) 0.00000  0.0000
+# varietySNMS:n.trtLATE SPRING                -20.06813  7867.38080
+# varietySNMS:status2CNTL                     -19.95030  7561.42603
+# varietySNMS:n.trtEARLY SPRING:status2CNTL    20.91952  7561.42612
+# varietySNMS:n.trtLATE SPRING:status2CNTL     20.67575 13394.17908
+# deviance df.resid
+#    726.8      147
+# 726.8 / 147 = 4.9442177
+
+glmmPlot <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|plot),
+                  control = glmerControl(optimizer = "bobyqa",
+                                         optCtrl = list("maxfun" = 20000)),
+                  family = binomial, data = fert2)
+# Warning message:
+# In checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv,  :
+#   Model is nearly unidentifiable: large eigenvalue ratio
+#  - Rescale variables?
+# plot   (Intercept) 1.3056   1.1426
+# varietySNMS:n.trtLATE SPRING              -19.20256  371.85490
+# varietySNMS:status2CNTL                   -19.21857  211.30941
+# varietySNMS:n.trtEARLY SPRING:status2CNTL  20.44165  211.30904
+# varietySNMS:n.trtLATE SPRING:status2CNTL   19.91369  260.18702
+# deviance df.resid
+#    731.7      149
+# 731.7 / 149 = 4.9107383
+
+glmmVarS <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
+                    (variety|block),
+                  control = glmerControl(optimizer = "bobyqa",
+                                         optCtrl = list("maxfun" = 20000)),
+                  family = binomial, data = fert2)
+# Warning messages:
+# 1: In checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv,  :
+#   unable to evaluate scaled gradient
+# 2: In checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv,  :
+#   Model failed to converge: degenerate  Hessian with 3 negative eigenvalues
+# block  (Intercept) 0.62664  0.79160
+#        varietyPRHN 0.98385  0.99189  -0.933
+#        varietySNMS 3.09082  1.75807  -0.713  0.876
+#        varietyTAM  0.38672  0.62186  -0.471  0.432  0.162
+#        varietyYSTN 0.86358  0.92929  -0.798  0.687  0.277  0.806
+# varietySNMS:n.trtLATE SPRING               -18.54752 2546.52186
+# varietySNMS:status2CNTL                    -18.51916 2537.53981
+# varietySNMS:n.trtEARLY SPRING:status2CNTL   19.13402 2537.53988
+# varietySNMS:n.trtLATE SPRING:status2CNTL    19.08471 4406.80309
+# deviance df.resid
+#    842.8      135
+# 842.8 / 135 = 6.2422222
+
+# Compare coefs and SEs
+cbind(coef(glmQBin)[c(1,7:13,34:55)], fixef(glmmVar))
+cbind(arm::se.coef(glmQBin)[c(1,7:13,34:55)], arm::se.fixef(glmmVar))
+# Most of the quasibinomial SEs are less than twice as big as the glmm SEs
+
+# Now with plot-level REs
+cbind(coef(glmQBin)[c(1,7:13,34:55)], fixef(glmmVar))
+cbind(arm::se.coef(glmQBin)[c(1,7:13,34:55)], arm::se.fixef(glmmVar))
+# Pretty similar. Maybe this is ok?
 @

### 2015-10-19 leslieandkenny.Rnw
Did some reorganizing, will work more out on paper and fill things in later

@@
 
-\subsection{Binomial Generalized Linear Model}
+\subsection{Potential Issues with Count Data}
 
-\noindent Because the response variable is a Binomial count, and because
-the design includes random assignment of variety to rows within blocks,
-a Binomial generalized linear model is appropriate. We recommend using
-a three-way interaction to estimate the probability of infection for each
-variety/treatment/inoculation status combination and a block/variety
-interaction to control for differences between blocks and between rows
-in the same block.\\
+\begin{figure}[h!]
+\begin{center}
+\caption{Illustration of separation.}
+<<separation,echo=FALSE,out.width='0.6\\linewidth',fig.width=6>>=
+curve(exp(x)/(1+exp(x)), from = -10, to = 10, ylim = c(0, 1),
+      xlab = expression(logit(p[i])), ylab = expression(p[i]))
+segments(x0 = -10, x1 = log(1/59), y0 = 1/60, lty = 2, lwd = 2)
+segments(x0 = log(1/59), y0 = 0, y1 = 1/60, lty = 2, lwd = 2)
+@
+\end{center}
+\end{figure}
 
-\noindent The model is
-$$
-y_i\sim\mathrm{overdispersed Binomial}(30, p_i, \omega),
-$$
-$$
-\mathrm{logit}(p_i)=\mu+\alpha_{b[i]}+\beta_{v[i]}+\gamma_{t[i]}+\delta_{j[i]}
-+(\alpha\beta)_{b[i],v[i]}+(\beta\gamma)_{v[i],t[i]}+(\beta\delta)_{v[i],j[i]}
-+(\gamma\delta)_{t[i],j[i]}+(\beta\gamma\delta)_{v[i],t[i],j[i]}
-$$
-where
-\begin{itemize}
-\item \(p_i\) is the probability of infection in the \(i\)th plot,
-\item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
-natural logarithm of the odds of infection, and
-\item \(b[i]\), \(v[i]\), \(t[i]\), and \(j[i]\) index the block,
-variety, nitrogen application timing, and inoculation status of
-observation \(i\),
-\item \(\omega\) is an overdispersion parameter, so that the standard
-deviation of the response in plot \(i\) is \(\sqrt{30\omega p_i(1-p_i)}\).
-\end{itemize}
-
-\noindent The assumptions for this model are that plants within a plot
-are independent and have the same probability of being infected, and
-that plots are independent given the other variables in the model. \\
-
-\noindent There are several reasons why we recommend using a fixed-effects
-generalized linear model instead of a generalized linear mixed model;
-these reasons are related to overdispersion and computational issues. \\
-
-\noindent Since the inoculated plants were planted in the center of the
-plots, we expect to see spatial dependence where plants near the centers
-are more likely to become infected than plants further from the centers.
-Then individual plants could have infection probabilities that differ from
-the infection rate estimated for the plot they are in. This would result
-in the data could having more variation than what the Binomial model
-accounts for, a situation known as overdispersion. We observed
-overdispersion in Nar's data, and there are no simple tools
-for including overdispersion in mixed models. Therefore, the standard
-errors in the mixed model output would be based on an incorrect estimate
-of the variance and might not be trustworthy. This problem is easily
-remedied by using a Quasibinomial fixed-effects GLM model. \\
-
-\noindent Additionally, including many coefficients and nested effects
-in a mixed model can cause computational problems where the software will
-fail to find unique solutions for the parameters. These problems are made
-worse in a Binomial GLMM when the counts are small, as in these data. The
-Binomial GLM can avoid many of the computational difficulties associated
-with the GLMM. The GLM can correctly model the nested design through a
-block by variety interaction. \\
-
-\noindent The only advantage of the GLMM is that it would estimate block-
-and row-level variances, but it is our opinion that this is outweighed by
-the lack of an overdispersion estimate and the computational difficulties.
+\noindent The following sections present examples of issues that could
+arise while analyzing Nar's data. The examples all use simulated datasets.
 
@@
 be considered as coming from a different population than the data with
-fewer zeros. \textbf{If and only if it is justified}, Nar could omit all
+fewer zeros. \emph{If and only if it is justified}, Nar could omit all
 of the data for this treatment or variety when fitting the model. The scope
@@
 <<tabplot,echo=FALSE,warning=FALSE,message=FALSE>>=
-#Warning message:
-#In tableplot_checkCols(substitute(sortCol), colNames) :
-#  Sorting on multiple columns not supported anymore. Only first column is taken.
-tableplot(fert.sim1, select = c(infected, variety, block, n.trt, status),
-          nBins = 30, sortCol = "variety")
+# Tableplot is confusing. Maybe omit this?
+fert.sim1$v.trt <- with(fert.sim1, interaction(status2, n.trt, variety))
+tableplot(fert.sim1, select = c(infected, variety, block, n.trt, status, v.trt),
+          nBins = 30, sortCol = "v.trt")
 @
@@
 
-\subsection{Model Fitting}
+\subsection{Binomial Generalized Linear Model}
+
+\noindent Because the response variable is a Binomial count, and because
+the design includes random assignment of variety to rows within blocks,
+a Binomial generalized linear model is appropriate. We recommend using
+a three-way interaction to estimate the probability of infection for each
+variety/treatment/inoculation status combination and a block/variety
+interaction to control for differences between blocks and between rows
+in the same block.\\
+
+\noindent The model is
+$$
+y_i\sim\text{overdispersed Binomial}(30, p_i, \omega),
+$$
+$$
+\mathrm{logit}(p_i)=\mu+\alpha_{b[i]}+\beta_{v[i]}+\gamma_{t[i]}+\delta_{j[i]}
++(\alpha\beta)_{b[i],v[i]}+(\beta\gamma)_{v[i],t[i]}+(\beta\delta)_{v[i],j[i]}
++(\gamma\delta)_{t[i],j[i]}+(\beta\gamma\delta)_{v[i],t[i],j[i]}
+$$
+where
+\begin{itemize}
+\item \(p_i\) is the probability of infection in the \(i\)th plot,
+\item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
+natural logarithm of the odds of infection, and
+\item \(b[i]\), \(v[i]\), \(t[i]\), and \(j[i]\) index the block,
+variety, nitrogen application timing, and inoculation status of
+observation \(i\),
+\item \(\omega\) is an overdispersion parameter, so that the variance
+of the response in plot \(i\) is \(30\omega p_i(1-p_i)\).
+\end{itemize}
+
+\noindent The assumptions for this model are that plants within a plot
+are independent and have the same probability of being infected, and
+that plots are independent given the other variables in the model. \\
+
+\noindent There are several reasons why we recommend using a fixed-effects
+generalized linear model instead of a generalized linear mixed model;
+these reasons are related to overdispersion and computational issues. \\
+
+\noindent Since the inoculated plants were planted in the center of the
+plots, we expect to see spatial dependence where plants near the centers
+are more likely to become infected than plants further from the centers.
+Then individual plants could have infection probabilities that differ from
+the infection rate estimated for the plot they are in. This would result
+in the data could having more variation than what the Binomial model
+accounts for, a situation known as overdispersion. We observed
+overdispersion in Nar's data, and there are no simple tools
+for including overdispersion in mixed models. Therefore, the standard
+errors in the mixed model output would be based on an incorrect estimate
+of the variance and might not be trustworthy. This problem is easily
+remedied by using a Quasibinomial fixed-effects GLM model. \\
+
+\noindent Additionally, including many coefficients and nested effects
+in a mixed model can cause computational problems where the software will
+fail to find unique solutions for the parameters. These problems are made
+worse in a Binomial GLMM when the counts are small, as in these data. The
+Binomial GLM can avoid many of the computational difficulties associated
+with the GLMM. The GLM can correctly model the nested design through a
+block by variety interaction. \\
+
+\noindent The only advantage of the GLMM is that it would estimate block-
+and row-level variances, but it is our opinion that this is outweighed by
+the lack of an overdispersion estimate and the computational difficulties.
+
+\subsection{Binomial Generalized Linear Mixed Model}
+\label{model2}
+
+\noindent The model is
+$$
+y_i\sim\mathrm{Binomial}(30, p_i),
+$$
+\begin{align*}
+\mathrm{logit}(p_i)&=\mu+a_{b[i]}+(ab)_{b[i],v[i]}\\
+&\quad+\beta_{v[i]}+\gamma_{t[i]}+\delta_{j[i]}\\
+&\quad+(\alpha\beta)_{b[i],v[i]}+(\beta\gamma)_{v[i],t[i]}
++(\beta\delta)_{v[i],j[i]}+(\gamma\delta)_{t[i],j[i]}\\
+&\quad+(\beta\gamma\delta)_{v[i],t[i],j[i]}
+\end{align*}
+where
+\begin{itemize}
+\item \(p_i\) is the probability of infection in the \(i\)th plot,
+\item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
+natural logarithm of the odds of infection, and
+\item \(b[i]\), \(v[i]\), \(t[i]\), and \(j[i]\) index the block,
+variety, nitrogen application timing, and inoculation status of
+observation \(i\),
+\item \(a_b\sim\mathrm{N}(0,\sigma^2_a)\) and
+\((ab)_{b,v}\sim\mathrm{N}(0,\sigma^2_{ab})\) are random adjustments for
+each block and each variety within each block.
+\end{itemize}
+
+\subsection{Fitting the GLM}
 \label{model}
@@
 <<glmms,echo=FALSE,warning=FALSE,message=FALSE,results='hide',cache=TRUE>>=
+# Experimenting with different random effects.
+# Are all block coefs 0 because they are nonidentified?
+
 glmBin <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
@@
 # Warning messages:
-# 1: In vcov.merMod(object, use.hessian = use.hessian) :
-#   variance-covariance matrix computed from finite-difference Hessian is
-# not positive definite or contains NA values: falling back to var-cov estimated from RX
-# 2: In vcov.merMod(object, correlation = correlation, sigm = sig) :
-#   variance-covariance matrix computed from finite-difference Hessian is
-# not positive definite or contains NA values: falling back to var-cov estimated from RX
-# plot          (Intercept) 1.03818  1.0189
+# 1: In checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv,  :
+#   unable to evaluate scaled gradient
+# 2: In checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv,  :
+#   Model failed to converge: degenerate  Hessian with 2 negative eigenvalues# plot          (Intercept) 1.03818  1.0189
 # variety:block (Intercept) 0.28324  0.5322
@@
 # Pretty similar. Maybe this is ok?
+# Mark doesn't like it.
 @
@@
 @
+
+\subsection{Fitting the GLMM}
 

### 2015-10-20 leslieandkenny.Rnw
revised model refinement section

@@
 #for leslie
-#setwd("~/Documents/ConsultingFall2015/Fall510")
+setwd("~/Documents/ConsultingFall2015/Fall510")
 fert2 <- read.csv("fert2.csv")
@@
 For example, the \verb+n.trt:status2+ row in the table tests for removal of the $n.trt*status2$ two-way interaction, conditional upon all other two way interactions and main effects being in the model.
-As a result, this table can be used to assess evidence for removal of each term.
+These tables can be used to assess all the effects at the same level together.
+If you choose to remove a term, however, it is important to fit a new model and reconsider the Anova table every time a simpler model is selected.
+This is necessary because the estimate of overdispersion changes when the model changes, causing the tests in the table to change as well.
+\\
 
-<<refine, echo=TRUE>>=
+\noindent For example, if we drop the $3$ way interaction term, the model with all the two-way interactions is fit and the Anova table is considered.
+
+<<anovadisplay, echo=TRUE, results='hide'>>=
 Anova(glm3way, test="F")
@@
 
-\noindent Here is how we would interpret the results in each row of the table.
+<<refine, echo=FALSE, results='asis'>>=
+print(xtable(Anova(glm3way, test="F"), caption="Anova table for the $3$ way interaction model."), caption.placement = "top")
+@
+
+<<anovadisplay2, echo=TRUE, results='hide'>>=
+glm2way <- glm(cbind(infected, total) ~ block*variety + variety*n.trt 
+               + variety*status2+n.trt*status2,
+               family = quasibinomial, data = fert.sim1)
+Anova(glm2way, test="F")
+@
+
+<<refitno3way, echo=FALSE, results='asis'>>=
+print(xtable(Anova(glm2way, test="F"), caption="Anova table for the model with all $2$ way interactions."), caption.placement = "top")
+@
+
+\noindent Here is how we would interpret the results in the rows of interest in the above tables.
 Note, it is not appropriate to consider removal of any of the terms involving block.
@@
 \item \verb+variety:n.trt:status2+ ``There is no evidence that the interaction between variety and nitrogen application time differs between inoculated and control plots (p-value$=0.388$ from F-stat$=1.07$ on $8$ and $125$ df).''
-\item \verb+n.trt:status2+ ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection differs between inoculated and control plots after controlling for all other two way interactions in the model (p-value$=0.368$ from F-stat$=1.01$ on $2$ and $133$ df).
-\item \verb+variety:status2+ ``There is no evidence that the relationship between variety and the odds of virus infection differs between inoculated and control plots after controlling for all other two way interactions in the model (p-value$=0.288$ from F-stat$=1.26$ on $4$ and $133$ df).
-\item \verb+variety:n.trt+ ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection depends on variety after controlling for all other two way interactions in the model (p-value$=0.453$ from F-stat$=0.98$ on $8$ and $133$ df).
+\item \verb+n.trt:status2+ ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection differs between inoculated and control plots after controlling for all other two way interactions and main effects in the model (p-value$=0.380$ from F-stat$=0.973$ on $2$ and $133$ df).
+\item \verb+variety:status2+ ``There is no evidence that the relationship between variety and the odds of virus infection differs between inoculated and control plots after controlling for all other two way interactions and main effects in the model (p-value$=0.304$ from F-stat$=1.22$ on $4$ and $133$ df).
+\item \verb+variety:n.trt+ ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection depends on variety after controlling for all other two way interactions and main effects in the model (p-value$=0.478$ from F-stat$=0.950$ on $8$ and $133$ df).
 \end{enumerate}
+
+\noindent If you remove another term from the model, refit the simpler model and reassess evidence for each effect.
+For example, suppose we remove the variety by inoculation status interaction term.
+Our assessment of the variety by nitrogen application interaction effect in the following simpler model would be, ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection depends on variety after controlling for all main effects and the two way interaction between variety and inoculation status (p-value$=0.4536$ from F-stat$=0.98$ on $8$ and $135$ df).
+
+<<refitno2way, echo=FALSE, results='asis'>>=
+glm2way2 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt 
+               + variety*status2,
+               family = quasibinomial, data = fert.sim1)
+print(xtable(Anova(glm2way2, test="F"), caption="Anova table for the model with two way interactions between variety and nitrogen and variety and inoculation status."), caption.placement = "top")
+@
+
+<<refitno2way3, echo=FALSE, results='asis', include=FALSE>>=
+glm2way3 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt,
+               family = quasibinomial, data = fert.sim1)
+print(xtable(Anova(glm2way3, test="F"), caption="Anova table for the model with a two way interaction between variety and nitrogen application time."), caption.placement = "top")
+@
 
@@
 
-<<glmnointeractions, echo=TRUE>>=
+<<glmnointeractions, echo=TRUE, results='hide'>>=
 glmnointeractions <- glm(cbind(infected, total) ~ block*variety + variety + n.trt + status2,
@@
 Anova(glmnointeractions, test="F")
+@
+
+<<glmnointeractions2, echo=FALSE, results='asis'>>=
+print(xtable(Anova(glmnointeractions, test="F"), caption="Anova table for the model with no interactions."), caption.placement="top")
 @

### 2015-10-20 leslieandkenny.Rnw
More reorganization

@@
 for spatial patterns. If correlation between nearby plots is seen,
-the model should be modified to account for this. \\
-
-\noindent Another option for examining the data graphically is the
-\verb|itableplot| function in the \verb|tabplot| package. This is a
-useful interactive tool for visualizing large datasets and identifying
-patterns.
-
-\begin{figure}[h!]
-\caption{Example \texttt{itableplot} output.}
-<<tabplot,echo=FALSE,warning=FALSE,message=FALSE>>=
-# Tableplot is confusing. Maybe omit this?
-fert.sim1$v.trt <- with(fert.sim1, interaction(status2, n.trt, variety))
-tableplot(fert.sim1, select = c(infected, variety, block, n.trt, status, v.trt),
-          nBins = 30, sortCol = "v.trt")
-@
-\end{figure}
+the model should be modified to account for this.
 
@@
 \end{itemize}
+
+\subsection{Comparison of the GLM and the GLMM}
 
@@
 # Now with plot-level REs
-cbind(coef(glmQBin)[c(1,7:13,34:55)], fixef(glmmVar))
-cbind(arm::se.coef(glmQBin)[c(1,7:13,34:55)], arm::se.fixef(glmmVar))
+cbind(coef(glmQBin)[c(1,7:13,34:55)], fixef(glmmVarPlot))
+cbind(arm::se.coef(glmQBin)[c(1,7:13,34:55)], arm::se.fixef(glmmVarPlot))
 # Pretty similar. Maybe this is ok?

### 2015-10-21 leslieandkenny.Rnw
added to comparison section

@@
 
-We address the following questions in this report:
+We address the following questions in this report. The first three questions are related to model selection, and the last question is related to interpretation after a model is selected.
 \begin{itemize}
@@
 
-\noindent There are several reasons why we recommend using a fixed-effects
-generalized linear model instead of a generalized linear mixed model;
-these reasons are related to overdispersion and computational issues. \\
-
-\noindent Since the inoculated plants were planted in the center of the
-plots, we expect to see spatial dependence where plants near the centers
-are more likely to become infected than plants further from the centers.
-Then individual plants could have infection probabilities that differ from
-the infection rate estimated for the plot they are in. This would result
-in the data could having more variation than what the Binomial model
-accounts for, a situation known as overdispersion. We observed
-overdispersion in Nar's data, and there are no simple tools
-for including overdispersion in mixed models. Therefore, the standard
-errors in the mixed model output would be based on an incorrect estimate
-of the variance and might not be trustworthy. This problem is easily
-remedied by using a Quasibinomial fixed-effects GLM model. \\
-
-\noindent Additionally, including many coefficients and nested effects
-in a mixed model can cause computational problems where the software will
-fail to find unique solutions for the parameters. These problems are made
-worse in a Binomial GLMM when the counts are small, as in these data. The
-Binomial GLM can avoid many of the computational difficulties associated
-with the GLMM. The GLM can correctly model the nested design through a
-block by variety interaction. \\
-
-\noindent The only advantage of the GLMM is that it would estimate block-
-and row-level variances, but it is our opinion that this is outweighed by
-the lack of an overdispersion estimate and the computational difficulties.
-
 \subsection{Binomial Generalized Linear Mixed Model}
@@
 
-\subsection{Comparison of the GLM and the GLMM}
+\subsection{Advantage of the Quasibinomial Generalized Linear Model over the Binomial Generalized Linear Mixed Model}
+
+\noindent There are several reasons why we recommend using a fixed-effects
+generalized linear model instead of a generalized linear mixed model in this report. These reasons are related to overdispersion and computational issues. \\
+
+\noindent Since the inoculated plants were planted in the center of the
+plots, we expect to see spatial dependence where plants near the centers
+are more likely to become infected than plants further from the center. 
+Then individual plants could have infection probabilities that differ from
+the infection rate estimated for the plot they are in. This would result
+in the data could having more variation than what the Binomial model
+accounts for, a situation known as overdispersion. We observed
+overdispersion in Nar's data, and there are no simple tools
+for including overdispersion in mixed models. Therefore, the standard
+errors in the mixed model output would be based on an incorrect estimate
+of the variance and might not be trustworthy.
+The standard errors in the GLMM model would be smaller than they should be, causing the p-values for effects of interest to be artificially small.
+This problem is easily remedied by using a quasibinomial fixed-effects GLM model.
+\\
+
+\noindent We illustrate this problem in the output below.
+We fit a quasibinomial logistic regression model, and a binomial generalized linear mixed effects model to the real data sent to us by Nar (details on model fitting in section \ref{model} and \ref{glmmfit}).
+The coefficients, standard errors, and tests are shown for the three way interaction coefficients in each model.
+The estimates are similar between the two models, but the standard errors are much larger in the quasibinomial GLM model.
+Without accounting for overdispersion, the standard errors for the estimates in the mixed effects model are artificially small.
+For this reason, we advise Nar not to trust the results from the mixed effects model.
+
+<<comparison, echo=FALSE, message=FALSE, results='asis'>>=
+glmQBin.noSNMS <- glm(cbind(infected, total) ~ block*variety + 
+                        variety*n.trt*status2,
+               family = quasibinomial, data = subset(fert2, variety != "SNMS"))
+print(xtable(summary(glmQBin.noSNMS)$coefficients[39:44,], caption="Estimates, standard errors, and tests for three way interaction coefficients in the quasibinomial model.", label="qbin"), caption.placement="top", size="footnotesize", table.placement = getOption("xtable.table.placement", "H"))
+
+noSNMS3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
+                    (1|block/row), subset = variety != "SNMS",
+                    control = glmerControl(optimizer = "bobyqa"),
+                    family = binomial, data = fert2)
+
+print(xtable(summary(noSNMS3way)$coefficients[19:24,], caption="Estimates, standard errors, and tests for three way interaction coefficients in the GLMER model", label="glmer"), caption.placement="top", size="footnotesize", table.placement = getOption("xtable.table.placement", "H"))
+@
+
+\noindent Additionally, including many coefficients and nested effects
+in a mixed model can cause computational problems where the software will
+fail to find unique solutions for the parameters. These problems are made
+worse in a binomial GLMM when the counts are small, as in these data. The
+binomial GLM can avoid many of the computational difficulties associated
+with the GLMM.  \\
+
+\noindent Rows are nested within blocks in the study design, and each row is randomly assigned to a variety.
+The design's nesting structure can be accounted for in a GLMM model  by including random effects for row and block.
+Originally, however, random effects were developed to reflect a feature of the study design.
+For example, if $10$ operators were chosen at random from a larger population of operators, {\it operator} would be modeled as a random effect.
+In this case, the sample of operators truly is random in the design, and the operators selected are intended to represent a larger population of operators.
+In Nar's study design, row and block are not truly random effects because they were not selected from a larger population of rows and blocks.
+It is important to recognize this and understand that the only reason row and block are included as random effects in the GLMM model is to account for the nesting structure of row within block.
+As a result, we think it is more appropriate to model the nested design using fixed effects in the binomial generalized linear model, and this strategy is more consistent with the study design.
+\\
+
+\noindent The binomial GLM can correctly model the nested design through a block by variety interaction.
+In the GLM model, it is possible to include a \verb+block/variety+ term, and this is equivalent to including the \verb+block*variety+ interaction.
+Including this interaction accounts for the nesting because it allows the differences in varieties to vary by block.
+\\
+
+\noindent The main advantages of the GLMM is that it would estimate block-
+and row-level variances, and it would require fewer degrees of freedom to account for the nesting structure in the design.
+It is our opinion that this is outweighed by the lack of an overdispersion estimate and the computational difficulties.
 
@@
 \subsection{Fitting the GLMM}
+\label{glmmfit}
 
@@
 
-\noindent For example, if we drop the $3$ way interaction term, the model with all the two-way interactions is fit and the Anova table is considered.
+\noindent The Anova table for the three-way interaction model is shown below (Table \ref{3way}).
+This table should only be used to assess evidence for the three-way interaction term.
+We recommend the following wording,
+\begin{quote}
+There is no evidence that the interaction between variety and nitrogen application time differs between inoculated and control plots (p-value$=0.388$ from F-stat$=1.07$ on $8$ and $125$ df).
+\end{quote}
 
@@
 <<refine, echo=FALSE, results='asis'>>=
-print(xtable(Anova(glm3way, test="F"), caption="Anova table for the $3$ way interaction model."), caption.placement = "top")
+print(xtable(Anova(glm3way, test="F"), caption="Anova table for the $3$ way interaction model.", label="3way"), caption.placement = "top", size='footnotesize', table.placement = getOption("xtable.table.placement", "H"))
 @
+
+\noindent If you drop the $3$ way interaction term, the model with all the two-way interactions should be fit and the Anova table analysis should be repeated.
 
@@
 <<refitno3way, echo=FALSE, results='asis'>>=
-print(xtable(Anova(glm2way, test="F"), caption="Anova table for the model with all $2$ way interactions."), caption.placement = "top")
+print(xtable(Anova(glm2way, test="F"), caption="Anova table for the model with all $2$ way interactions.", label="twoway"), caption.placement = "top", size='footnotesize', table.placement = getOption("xtable.table.placement", "H"))
 @
 
-\noindent Here is how we would interpret the results in the rows of interest in the above tables.
-Note, it is not appropriate to consider removal of any of the terms involving block.
-Block to block variability is expected because it was included as a design variable, and so it should be included in the model.
-The $block*variety$ interaction also needs to be included to account for the nesting structure in the design.
+\noindent Here is how we would interpret the results in the applicable rows in the above the two-way interaction Anova table (Table \ref{twoway}). 
 \begin{enumerate}
-\item \verb+variety:n.trt:status2+ ``There is no evidence that the interaction between variety and nitrogen application time differs between inoculated and control plots (p-value$=0.388$ from F-stat$=1.07$ on $8$ and $125$ df).''
 \item \verb+n.trt:status2+ ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection differs between inoculated and control plots after controlling for all other two way interactions and main effects in the model (p-value$=0.380$ from F-stat$=0.973$ on $2$ and $133$ df).
@@
 \noindent If you remove another term from the model, refit the simpler model and reassess evidence for each effect.
-For example, suppose we remove the variety by inoculation status interaction term.
-Our assessment of the variety by nitrogen application interaction effect in the following simpler model would be, ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection depends on variety after controlling for all main effects and the two way interaction between variety and inoculation status (p-value$=0.4536$ from F-stat$=0.98$ on $8$ and $135$ df).
-
+For example, suppose we remove the variety by inoculation status interaction term (Table \ref{remove}).
+Our assessment of the variety by nitrogen application interaction effect in the following simpler model would be,
+\begin{quote}
+There is no evidence that the relationship between nitrogen application time and the odds of virus infection depends on variety after controlling for all main effects and the two way interaction between variety and inoculation status (p-value$=0.4536$ from F-stat$=0.98$ on $8$ and $135$ df).
+\end{quote}
 <<refitno2way, echo=FALSE, results='asis'>>=
@@
                family = quasibinomial, data = fert.sim1)
-print(xtable(Anova(glm2way2, test="F"), caption="Anova table for the model with two way interactions between variety and nitrogen and variety and inoculation status."), caption.placement = "top")
+print(xtable(Anova(glm2way2, test="F"), caption="Anova table for the model with two way interactions between variety and nitrogen and variety and inoculation status.", label="remove"), caption.placement = "top", size='footnotesize', table.placement = getOption("xtable.table.placement", "H"))
 @
@@
                family = quasibinomial, data = fert.sim1)
-print(xtable(Anova(glm2way3, test="F"), caption="Anova table for the model with a two way interaction between variety and nitrogen application time."), caption.placement = "top")
+print(xtable(Anova(glm2way3, test="F"), caption="Anova table for the model with a two way interaction between variety and nitrogen application time."), caption.placement = "top", table.placement = getOption("xtable.table.placement", "H"))
 @
 
-\noindent If no evidence is found for any of the interactions, the simplest model we recommend fitting should contain the $block*variety$ interaction and all main effects (shown below).
-If this model is selected, each effect is then interpreted after controlling for the effects of the other variables in the model.
-For example, to assess evidence for the \verb+n.trt+ effect in this model, we would say, ``There is no evidence that the odds of virus infection depends on nitrogen application time after controlling for block, variety, inoculation status, and the
-$block*variety$ interaction.''
+\noindent If no evidence is found for any of the interactions, the simplest model we recommend fitting should contain the $block*variety$ interaction and all main effects (Table \ref{simple}).
+Note, it is not appropriate to consider removal of any of the terms involving block.
+Block to block variability is expected because it was included as a design variable, and so it should be included in the model.
+The $block*variety$ interaction also needs to be included to account for the nesting structure in the design.
+\\
+
+\noindent To assess evidence for the \verb+n.trt+ effect in the simplest recommended model, we would say, 
+\begin{quote}
+There is no evidence that the odds of virus infection depends on nitrogen application time after controlling for block, variety, inoculation status, and the $block*variety$ interaction (p-value$=0.612$ from F-stat$=0.49$ on $2$ and $147$ df).
+\end{quote}
 
@@
 <<glmnointeractions2, echo=FALSE, results='asis'>>=
-print(xtable(Anova(glmnointeractions, test="F"), caption="Anova table for the model with no interactions."), caption.placement="top")
+print(xtable(Anova(glmnointeractions, test="F"), caption="Anova table for the model with no interactions.", label="simple"), caption.placement="top", size='footnotesize', table.placement = getOption("xtable.table.placement", "H"))
 @

### 2015-10-21 leslieandkenny.Rnw
done with count issues

--- 2015-10-20 leslieandkenny.Rnw
@@
 #for leslie
-setwd("~/Documents/ConsultingFall2015/Fall510")
+#setwd("~/Documents/ConsultingFall2015/Fall510")
 fert2 <- read.csv("fert2.csv")
@@
 \section{Introduction}
-\noindent
 
-\section{Statistical Experience and Assistance Needed}
 \noindent Nar B.
@@
 
+The response variable is the count of infected leaves out of the 30 leaves
+collected in each plot. The standard approach to analyzing count data is
+to use a Binomial distribution and model the probability \(p_i\) that a leaf
+in a given plot \(i\) will get infected. This modeling occurs on the scale
+of the log-odds, \(\mathrm{logit}(p_i)=\log\left(\frac{p_i}{1-p_i}\right)\). \\
+
+\noindent One serious problem that can occur when working with counts is
+the issue of separation. This is when one variable or a combination of factors
+is perfectly associated with the response. For example, if a certain
+treatment results in an infection count of zero, then knowing that a leaf
+recieved that treatment determines that the leaf will not get infected,
+regardless of other variables. This causes difficulty in estimating the
+effect of that treatment on the infection rate, and results a large standard
+error because the effect cannot be precisely determined. \\
+
+\noindent Figure 1 provides a visual illustration of the separation problem.
+The left panel shows a small probability on the vertical axis and its
+corresponding logit value on the horizontal axis. The right panel zooms
+in on the lower left corner. The dotted lines show how a small amount
+of uncertainty on the probability scale translates to a relatively larger
+amount of uncertainty on the logit scale. \\
+
 \begin{figure}[h!]
@@
 \caption{Illustration of separation.}
-<<separation,echo=FALSE,out.width='0.6\\linewidth',fig.width=6>>=
+<<separation,echo=FALSE>>=
+par(mfrow = c(1, 2), mar = c(4, 4, 2, 2))
 curve(exp(x)/(1+exp(x)), from = -10, to = 10, ylim = c(0, 1),
@@
 segments(x0 = log(1/59), y0 = 0, y1 = 1/60, lty = 2, lwd = 2)
+
+curve(exp(x)/(1+exp(x)), from = -10, to = -2, ylim = c(0, 0.4),
+      xlab = expression(logit(p[i])), ylab = expression(p[i]))
+segments(x0 = -10, x1 = log(1/59), y0 = 1/60, lty = 2, lwd = 2)
+segments(x0 = log(1/59), y0 = 0, y1 = 1/60, lty = 2, lwd = 2)
+segments(x0 = -10, x1 = log(1/39), y0 = 1/40, lty = 3, lwd = 1)
+segments(x0 = log(1/39), y0 = 0, y1 = 1/40, lty = 3, lwd = 1)
+segments(x0 = -10, x1 = log(1/119), y0 = 1/120, lty = 3, lwd = 1)
+segments(x0 = log(1/119), y0 = 0, y1 = 1/120, lty = 3, lwd = 1)
 @
@@
 
+\noindent The problem gets worse if the infected count is 0. In that
+case, the best estimate of the infection probability is 0, and
+\(\mathrm{logit}(0)=-\infty\). This will cause computational problems
+because the software will try to improve the model fit by decreasing
+the effect estimate toward \(-\infty\). The software may fail to find a
+solution, and if it does report a solution, the standard error will be
+large because the solution cannot be found precisely. \\
+
 \noindent The following sections present examples of issues that could
-arise while analyzing Nar's data. The examples all use simulated datasets.
+arise while analyzing Nar's data, and our recommendations on dealing with
+them. The examples all use simulated datasets.
 
@@
 
-\noindent We noticed that infection counts of 0 were recorded in 67 of the
-180 plots in Nar's dataset. This is a large fraction of the data; if the
-distribution of these zeros is related to the variables in the model then
-it may be impossible to estimate coefficients for factor levels where
-infection rates are nearly zero. \\
-
-\noindent We recommend sorting the data and creating a heatmap to look
-for patterns of zeros. Figure 2 shows an example using simulated data.
-The most serious problems will come from the zeros, so we chose a contrasting
-color (black) for plots with infection counts of 0.\\
-
-\noindent Any treatment or variety with many black squares is a reason
-for concern. In the example, all but one of the inoculated fall-application
-plots for Var1 had counts of 0. The software will have difficulty estimating
-the the three-way interaction involving fall nitrogen application and
-variety 1. \\
-
-\noindent If a treatment has many zeros across all varieties or a variety
-has many zeros, Nar should consider whether he has reason to believe
-that treatment or variety has qualitative differences from the others
-that would cause it to have a much lower infection rate. These data would
-be considered as coming from a different population than the data with
-fewer zeros. \emph{If and only if it is justified}, Nar could omit all
-of the data for this treatment or variety when fitting the model. The scope
-of inference for this model must then be limited to varieties and
-treatments that have a nonzero infection rate. A separate analysis would
-be needed for the more resistant population. \\
+\noindent A graphical analysis is also a good way to see if separation
+is present in the data. In Nar's dataset, 67 of the 180 counts are
+0. This is a large fraction of the data, so it is very important to look
+for possible separation. We recommend sorting the data and creating a
+heatmap to look for patterns of zeros. Figure 2 shows an example using
+simulated data. The most serious problems will come from the zeros, so we
+chose a contrasting color (black) for plots with infection counts of 0. \\
 
@@
 
+\noindent Any treatment or variety with many black squares is a reason
+for concern. In the example, all but one of the inoculated fall-application
+plots for Var1 had counts of 0. The software will have difficulty estimating
+the the three-way interaction involving fall nitrogen application and
+variety 1. \\
+
+\noindent If a treatment has many zeros across all varieties or a variety
+has many zeros, Nar should consider whether he has reason to believe
+that treatment or variety has qualitative differences from the others
+that would cause it to have a much lower infection rate. These data would
+be considered as coming from a different population than the data with
+fewer zeros. \emph{If and only if it is justified}, Nar could omit all
+of the data for this treatment or variety when fitting the model. The scope
+of inference for this model must then be limited to varieties and
+treatments that have a nonzero infection rate. \\
+
+\noindent The graphic could be presented as evidence that the omitted
+data does not fit the same model as the data with higher infection counts.
+A separate analysis would be needed if additional information is desired
+about the more resistant population. \\
+
 \noindent If information about the physical layout of the field is
@@
 $$
-$$
-\mathrm{logit}(p_i)=\mu+\alpha_{b[i]}+\beta_{v[i]}+\gamma_{t[i]}+\delta_{j[i]}
-+(\alpha\beta)_{b[i],v[i]}+(\beta\gamma)_{v[i],t[i]}+(\beta\delta)_{v[i],j[i]}
-+(\gamma\delta)_{t[i],j[i]}+(\beta\gamma\delta)_{v[i],t[i],j[i]}
-$$
+\begin{align*}
+\mathrm{logit}(p_i)&=\mu+\alpha_{b[i]}+\beta_{v[i]}
++\gamma_{t[i]}+\delta_{j[i]}\\
+&\quad+(\alpha\beta)_{b[i],v[i]}+(\beta\gamma)_{v[i],t[i]}
++(\beta\delta)_{v[i],j[i]}+(\gamma\delta)_{t[i],j[i]}
++(\beta\gamma\delta)_{v[i],t[i],j[i]}
+\end{align*}
 where
@@
 \begin{align*}
-\mathrm{logit}(p_i)&=\mu+a_{b[i]}+(ab)_{b[i],v[i]}\\
-&\quad+\beta_{v[i]}+\gamma_{t[i]}+\delta_{j[i]}\\
+\mathrm{logit}(p_i)&=\mu+a_{b[i]}+(ab)_{b[i],v[i]}
++\beta_{v[i]}+\gamma_{t[i]}+\delta_{j[i]}\\
 &\quad+(\alpha\beta)_{b[i],v[i]}+(\beta\gamma)_{v[i],t[i]}
-+(\beta\delta)_{v[i],j[i]}+(\gamma\delta)_{t[i],j[i]}\\
-&\quad+(\beta\gamma\delta)_{v[i],t[i],j[i]}
++(\beta\delta)_{v[i],j[i]}+(\gamma\delta)_{t[i],j[i]}
++(\beta\gamma\delta)_{v[i],t[i],j[i]}
 \end{align*}

### 2015-10-22 leslieandkenny.Rnw
finished my sections, fixed typos and wording, I think this draft is ready to go

@@
 Nar has asked for our help in building a model, fitting it in R, and interpreting the results.
+In this report we describe the study and its potential issues, present our recommendations, and provide example R code.
 
@@
 
-We address the following questions in this report. The first three questions are related to model selection, and the last question is related to interpretation after a model is selected.
+\noindent We address the following questions in this report. The first three questions are related to model selection, and the last question is related to interpretation after a model is selected.
 \begin{itemize}
@@
 \noindent One serious problem that can occur when working with counts is
-the issue of separation. This is when one variable or a combination of factors
+separation. This is when one variable or a combination of factors
 is perfectly associated with the response. For example, if a certain
 treatment results in an infection count of zero, then knowing that a leaf
-recieved that treatment determines that the leaf will not get infected,
+received that treatment determines that the leaf will not get infected,
 regardless of other variables. This causes difficulty in estimating the
-effect of that treatment on the infection rate, and results a large standard
+effect of that treatment on the infection rate, and results in a large standard
 error because the effect cannot be precisely determined. \\
 
-\noindent Figure 1 provides a visual illustration of the separation problem.
-The left panel shows a small probability on the vertical axis and its
-corresponding logit value on the horizontal axis. The right panel zooms
-in on the lower left corner. The dotted lines show how a small amount
-of uncertainty on the probability scale translates to a relatively larger
-amount of uncertainty on the logit scale. \\
+\noindent Figure 1 provides a visual illustration of the
+separation problem. The left panel shows a small probability on the
+vertical axis and its corresponding logit value on the horizontal axis.
+The right panel zooms in on the lower left corner. The dotted lines show
+how a small amount of uncertainty on the probability scale translates to
+a relatively larger amount of uncertainty on the logit scale. \\
 
 \begin{figure}[h!]
+\label{sep}
 \begin{center}
 \caption{Illustration of separation.}
-<<separation,echo=FALSE>>=
+<<separation,echo=FALSE,out.height='0.3\\linewidth',fig.height=3>>=
 par(mfrow = c(1, 2), mar = c(4, 4, 2, 2))
@@
 arise while analyzing Nar's data, and our recommendations on dealing with
-them. The examples all use simulated datasets.
+them. Except where stated otherwise, the examples use simulated datasets.
 
@@
 \begin{figure}[h!]
+\label{qp}
 \caption{Example interaction plot using \texttt{qplot}.}
@@
 is present in the data. In Nar's dataset, 67 of the 180 counts are
-0. This is a large fraction of the data, so it is very important to look
-for possible separation. We recommend sorting the data and creating a
-heatmap to look for patterns of zeros. Figure 2 shows an example using
+0. This is a large fraction of the data, so it is very important to identify
+possible separation. We recommend sorting the data and creating a
+heatmap to look for patterns of zeros. Figure 3 shows an example using
 simulated data. The most serious problems will come from the zeros, so we
-chose a contrasting color (black) for plots with infection counts of 0. \\
+chose a contrasting color (black) for plots with infection counts of 0.
+The R code to produce this graphic appears in the appendix. \\
 
 \begin{figure}[h!]
+\label{hm}
 \caption{Example heatmap visualization for identifying patterns of zeros.}
@@
 be considered as coming from a different population than the data with
-fewer zeros. \emph{If and only if it is justified}, Nar could omit all
+fewer zeros. \textbf{If and only if it is justified}, Nar could omit all
 of the data for this treatment or variety when fitting the model. The scope
@@
 
-\subsection{Binomial Generalized Linear Model}
-
-\noindent Because the response variable is a Binomial count, and because
-the design includes random assignment of variety to rows within blocks,
-a Binomial generalized linear model is appropriate. We recommend using
-a three-way interaction to estimate the probability of infection for each
-variety/treatment/inoculation status combination and a block/variety
-interaction to control for differences between blocks and between rows
-in the same block.\\
+\subsection{Quasibinomial Generalized Linear Model}
+
+\noindent We recommend starting with a quasibinomial generalized linear
+model. This model includes a three-way interaction to estimate the
+infection probability for each variety/treatment/inoculation status
+combination, and a block/variety interaction to control for differences
+between blocks and between rows in the same block.\\
 
@@
 
+\noindent The overdispersion parameter accounts for slight violations
+of these assumptions. We expect the plants within a single plot to
+have different infection probabilities based on their proximity to
+plants where mites are present. This issue manifests itself in the form
+of excess variation because the probability that a particular plant
+gets infected will not be the same as the estimated probability for
+plants in its plot. The overdispersion is an estimate of this extra
+variability; without it the standard errors will be underestimated. \\
+
 \subsection{Binomial Generalized Linear Mixed Model}
 \label{model2}
+
+\noindent Another option is to use random effects to model the nested
+structure of row within block. This model includes a three-way
+interaction of fixed effects and does not include an overdispersion
+parameter. In mixed models, the random effects are meant to explain
+all the variation in the data, so overdispersed models are not
+used in this setting. \\
 
@@
 
+\noindent The assumptions for this model are that plants within a plot
+are independent and have the same probability of being infected, that
+blocks are independent, that rows in different blocks are independent, that
+plots in different rows are independent given the other variables in the
+model, and that there are no additional sources of variation besides
+block, row, and the Binomial distribution for the infection counts. \\
+
+\noindent This model accounts for block-to-block and row-to-row variation
+by including variance terms for the block and row random effects. The
+benefit of this is that the amount of variation described by the blocks
+and rows will be estimated. The drawback is that, if there is additional
+variation not explained by these random effects, the model will fit poorly.
+
 \subsection{Advantage of the Quasibinomial Generalized Linear Model over the Binomial Generalized Linear Mixed Model}
@@
 the infection rate estimated for the plot they are in. This would result
-in the data could having more variation than what the Binomial model
-accounts for, a situation known as overdispersion. We observed
-overdispersion in Nar's data, and there are no simple tools
-for including overdispersion in mixed models. Therefore, the standard
-errors in the mixed model output would be based on an incorrect estimate
-of the variance and might not be trustworthy.
+in the data having more variation than what the Binomial model accounts for,
+a situation known as overdispersion. We observed overdispersion in Nar's
+data, and mixed models have limited ability to account for overdispersion
+beyond what can be explained by the variables in the model. Therefore,
+the standard errors in the mixed model output would be based on an
+incorrect estimate of the variance and might not be trustworthy.
 The standard errors in the GLMM model would be smaller than they should be, causing the p-values for effects of interest to be artificially small.
@@
 The estimates are similar between the two models, but the standard errors are much larger in the quasibinomial GLM model.
-Without accounting for overdispersion, the standard errors for the estimates in the mixed effects model are artificially small.
+Without accounting for overdispersion, the standard errors for the estimates in the mixed effects model are too small.
 For this reason, we advise Nar not to trust the results from the mixed effects model.
@@
 \noindent Rows are nested within blocks in the study design, and each row is randomly assigned to a variety.
-The design's nesting structure can be accounted for in a GLMM model  by including random effects for row and block.
+The nesting structure of the design can be accounted for in a GLMM model  by including random effects for row and block.
 Originally, however, random effects were developed to reflect a feature of the study design.
@@
 
+\pagebreak
 \subsection{Fitting the GLM}
@@
 
-\noindent If it is decided that certain observations come from an extremely
-resistant population and can be omitted from the current analysis:
+\noindent The summary output should be examined for unexpectedly large
+estimates or standard errors. A warning message may or may not appear in
+this situation. If it is decided that certain observations come from an
+extremely resistant population and can be omitted from the current analysis:
 
@@
 @
+
+\subsection{Fitting the GLMM}
+\label{glmmfit}
+
+To fit the three-way interaction model:
+
+<<glmm1,eval=FALSE>>=
+glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/variety),
+                  control = glmerControl(optimizer = "bobyqa"),
+                  family = binomial, data = fert.sim1)
+@
+
+\noindent The \verb|control = glmerControl(optimizer = "bobyqa")| argument
+tells the \verb|glmer| function which numerical optimizer to use to find
+the estimates. Several optimizers are included with R, but \verb|bobyqa|
+works well on problems that are mathematically difficult. We recommend using
+it for all \verb|glmer| fits. \\
+
+\noindent The following warning message will probably appear:
+
+<<glmmhide1,echo=FALSE,cache=TRUE>>=
+glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/variety),
+                  control = glmerControl(optimizer = "bobyqa"),
+                  family = binomial, data = fert.sim1)
+@
+
+\noindent This message alone does not indicate any problems, but it
+appears because the model has a large number of parameters, which could
+cause the numerical optimizer to take many iterations to converge. By
+default, the algorithm will run for a maximum of 10,000 iterations. That
+number can be increased by specifying a larger \verb|maxfun| option:
+
+<<glmm2,eval=FALSE>>=
+glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/variety),
+                  control = glmerControl(optimizer = "bobyqa",
+                                         optCtrl = list("maxfun" = 20000)),
+                  family = binomial, data = fert.sim1)
+@
+
+\noindent If a warning message says something like ``maximum number of
+function evaluations exceeded'' or mentions ``max$\lvert$grad$\rvert$''
+and a tolerance level, then the optimizer ran for the full number of
+iterations without finding a solution. Increasing \verb|maxfun| as above
+may solve the problem. \\
+
+\noindent If messages about eigenvalues or ``unable to evaluate scaled
+gradient'' occur:
+
+<<glmmhide2,echo=FALSE,cache=TRUE>>=
+glmmPlot <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|plot),
+                  control = glmerControl(optimizer = "bobyqa",
+                                         optCtrl = list("maxfun" = 20000)),
+                  family = binomial, data = fert2)
+glmmVar <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
+                   (1|block/variety),
+                 control = glmerControl(optimizer = "bobyqa",
+                                        optCtrl = list("maxfun" = 20000)),
+                 family = binomial, data = fert2)
+@
+
+\noindent These indicate that unique coefficient estimates do not exist.
+This is most likely caused by separation, so the output should be
+examined for large standard errors. The best solution would be to fit
+a simpler model (such as one with fewer interaction terms) or to remove
+problematic variables or data. The decision to remove any terms or data
+must be justified by prior knowledge and not based on the R output.
+If this cannot be justified, then the generalized linear mixed model should
+not be used. \\
 
@@
 
-\subsection{Fitting the GLMM}
-\label{glmmfit}
-
 \subsection{Model Refinement}
@@
 In this section, we describe how to go through the process of model selection.
-As discussed in the previous section, we recommend starting with the most complicated three-way interaction model and then assessing which terms can be removed.
+As discussed in the previous sections, we recommend starting with the most complicated three-way interaction model and then assessing which terms can be removed.
 In general, start by considering removal of the highest order terms.
@@
 \noindent We'll start by assessing evidence for the three-way interaction term.
-One way to consider it's removal would be to fit a model without the three-way interaction, and then conduct a drop-in-deviance F-test comparing the full and reduced models, as you learned to do in Stat $512$.
+One way to consider its removal would be to fit a model without the three-way interaction, and then conduct a drop-in-deviance F-test comparing the full and reduced models, as you learned to do in Stat $512$.
 This method can be time consuming, so instead we recommend using the \verb+Anova+ function in the \verb+car+ package with the option \verb+test="F"+.
@@
 
+\pagebreak
 \section{References}
@@
 
-\noindent Sloderbeck, J., Michaud, P., Whitworth, Robert. "Wheat Pests." CurlMite. Kansas State University, 1 May 2008. Web. 18 Sept. 2015.\\
+\noindent Sloderbeck, J., Michaud, P., Whitworth, Robert. ``Wheat Pests.'' CurlMite. Kansas State University, 1 May 2008. Web. 18 Sept. 2015.\\
 {\footnotesize\verb|http://entomology.k-state.edu/extension/insect-information/crop-pests/wheat/curlmite.html|.}

### 2015-10-22 leslieandkenny.Rnw
added a sentence about fake data in Model Refinement Section

@@
 In this section, we describe how to go through the process of model selection.
-As discussed in the previous sections, we recommend starting with the most complicated three-way interaction model and then assessing which terms can be removed.
+As discussed in the previous section, we recommend starting with the most complicated  model and then assessing which terms can be removed.
 In general, start by considering removal of the highest order terms.
 \\
+
+\noindent Warning: the results reported in this section are from simulated data. This section is meant to be used as an example that can be followed when working with the real dataset. \\
 

### 2015-10-26 leslieandkenny.Rnw
Kenny is feeling humbled. Fixed model code and corrected interpretations to match.

@@
 #  "infected" = sample(fert2$infected, 180, replace = TRUE))
+fert.sim1$uninfected <- with(fert.sim1, total - infected)
 @
@@
 <<comparison, echo=FALSE, message=FALSE, results='asis'>>=
-glmQBin.noSNMS <- glm(cbind(infected, total) ~ block*variety +
+glmQBin.noSNMS <- glm(cbind(infected, uninfected) ~ block*variety +
                         variety*n.trt*status2,
@@
 
-noSNMS3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
+noSNMS3way <- glmer(cbind(infected, uninfected) ~ variety*n.trt*status2 +
                     (1|block/row), subset = variety != "SNMS",
@@
 <<startingmodel,warning=FALSE,results='hide',cache=TRUE>>=
-glm3way <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
+glm3way <- glm(cbind(infected, uninfected) ~ block*variety + variety*n.trt*status2,
                family = quasibinomial, data = fert.sim1)
@@
 <<novar1,warning=FALSE,results='hide',cache=TRUE>>=
-noVar1 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
+noVar1 <- glm(cbind(infected, uninfected) ~ block*variety + variety*n.trt*status2,
               family = quasibinomial, data = fert.sim1, subset = variety!="Var1")
@@
 <<separatemodels,warning=FALSE,results='hide',cache=TRUE>>=
-onlyINOC <- glm(cbind(infected, total) ~ block*variety + variety*n.trt,
+onlyINOC <- glm(cbind(infected, uninfected) ~ block*variety + variety*n.trt,
                 family = quasibinomial, data = fert.sim1, subset = status=="INOC")
 
-onlyCNTL <- glm(cbind(infected, total) ~ block*variety + variety*n.trt,
+onlyCNTL <- glm(cbind(infected, uninfected) ~ block*variety + variety*n.trt,
                 family = quasibinomial, data = fert.sim1, subset = status=="CNTL")
@@
 <<models5,echo=FALSE,warning=FALSE,message=FALSE,results='hide',cache=TRUE>>=
-glm3way2 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
+glm3way2 <- glm(cbind(infected, uninfected) ~ block*variety + variety*n.trt*status2,
                 family = quasibinomial, data = fert2)
@@
 
-glm2way2 <- glm(cbind(infected, total) ~ block*variety + (variety+n.trt+status2)^2,
+glm2way2 <- glm(cbind(infected, uninfected) ~ block*variety + (variety+n.trt+status2)^2,
                 family = quasibinomial, data = fert2)
@@
 <<glmm1,eval=FALSE>>=
-glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/variety),
+glmm3way <- glmer(cbind(infected, uninfected) ~ variety*n.trt*status2 + (1|block/variety),
                   control = glmerControl(optimizer = "bobyqa"),
@@
 <<glmmhide1,echo=FALSE,cache=TRUE>>=
-glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/variety),
+glmm3way <- glmer(cbind(infected, uninfected) ~ variety*n.trt*status2 + (1|block/variety),
                   control = glmerControl(optimizer = "bobyqa"),
@@
 <<glmm2,eval=FALSE>>=
-glmm3way <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|block/variety),
+glmm3way <- glmer(cbind(infected, uninfected) ~ variety*n.trt*status2 + (1|block/variety),
                   control = glmerControl(optimizer = "bobyqa",
@@
 <<glmmhide2,echo=FALSE,cache=TRUE>>=
-glmmPlot <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|plot),
+glmmPlot <- glmer(cbind(infected, uninfected) ~ variety*n.trt*status2 + (1|plot),
                   control = glmerControl(optimizer = "bobyqa",
@@
                   family = binomial, data = fert2)
-glmmVar <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
+glmmVar <- glmer(cbind(infected, uninfected) ~ variety*n.trt*status2 +
                    (1|block/variety),
@@
 
-glmBin <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
+glmBin <- glm(cbind(infected, uninfected) ~ block*variety + variety*n.trt*status2,
               family = binomial, data = fert2)
@@
 
-glmQBin <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
+glmQBin <- glm(cbind(infected, uninfected) ~ block*variety + variety*n.trt*status2,
                family = quasibinomial, data = fert2)
@@
 
-glmmVar <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
+glmmVar <- glmer(cbind(infected, uninfected) ~ variety*n.trt*status2 +
                    (1|block/variety),
@@
 
-glmmVarPlot <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
+glmmVarPlot <- glmer(cbind(infected, uninfected) ~ variety*n.trt*status2 +
                        (1|block/variety) + (1|plot),
@@
 
-glmmRow <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
+glmmRow <- glmer(cbind(infected, uninfected) ~ variety*n.trt*status2 +
                    (1|block/row),
@@
 
-glmmRowPlot <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
+glmmRowPlot <- glmer(cbind(infected, uninfected) ~ variety*n.trt*status2 +
                        (1|block/row) + (1|plot),
@@
 
-glmmPlot <- glmer(cbind(infected, total) ~ variety*n.trt*status2 + (1|plot),
+glmmPlot <- glmer(cbind(infected, uninfected) ~ variety*n.trt*status2 + (1|plot),
                   control = glmerControl(optimizer = "bobyqa",
@@
 
-glmmVarS <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
+glmmVarS <- glmer(cbind(infected, uninfected) ~ variety*n.trt*status2 +
                     (variety|block),
@@
 
-noSNMS2way <- glmer(cbind(infected, total) ~ variety*n.trt + variety*status2 +
+noSNMS2way <- glmer(cbind(infected, uninfected) ~ variety*n.trt + variety*status2 +
                     n.trt*status2 + (1|block/row), subset = variety != "SNMS",
@@
 \begin{quote}
-There is no evidence that the interaction between variety and nitrogen application time differs between inoculated and control plots (p-value$=0.388$ from F-stat$=1.07$ on $8$ and $125$ df).
+There is no evidence that the interaction between variety and nitrogen application time differs between inoculated and control plots (p-value$=0.406$ from F-stat$=1.05$ on $8$ and $125$ df).
 \end{quote}
@@
 <<anovadisplay2, echo=TRUE, results='hide'>>=
-glm2way <- glm(cbind(infected, total) ~ block*variety + variety*n.trt
+glm2way <- glm(cbind(infected, uninfected) ~ block*variety + variety*n.trt
                + variety*status2+n.trt*status2,
@@
 \begin{enumerate}
-\item \verb+n.trt:status2+ ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection differs between inoculated and control plots after controlling for all other two way interactions and main effects in the model (p-value$=0.380$ from F-stat$=0.973$ on $2$ and $133$ df).
-\item \verb+variety:status2+ ``There is no evidence that the relationship between variety and the odds of virus infection differs between inoculated and control plots after controlling for all other two way interactions and main effects in the model (p-value$=0.304$ from F-stat$=1.22$ on $4$ and $133$ df).
-\item \verb+variety:n.trt+ ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection depends on variety after controlling for all other two way interactions and main effects in the model (p-value$=0.478$ from F-stat$=0.950$ on $8$ and $133$ df).
+\item \verb+n.trt:status2+ ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection differs between inoculated and control plots after controlling for all other two way interactions and main effects in the model (p-value$=0.386$ from F-stat$=0.959$ on $2$ and $133$ df).
+\item \verb+variety:status2+ ``There is no evidence that the relationship between variety and the odds of virus infection differs between inoculated and control plots after controlling for all other two way interactions and main effects in the model (p-value$=0.302$ from F-stat$=1.23$ on $4$ and $133$ df).
+\item \verb+variety:n.trt+ ``There is no evidence that the relationship between nitrogen application time and the odds of virus infection depends on variety after controlling for all other two way interactions and main effects in the model (p-value$=0.465$ from F-stat$=0.967$ on $8$ and $133$ df).
 \end{enumerate}
@@
 \begin{quote}
-There is no evidence that the relationship between nitrogen application time and the odds of virus infection depends on variety after controlling for all main effects and the two way interaction between variety and inoculation status (p-value$=0.4536$ from F-stat$=0.98$ on $8$ and $135$ df).
+There is no evidence that the relationship between nitrogen application time and the odds of virus infection depends on variety after controlling for all main effects and the two way interaction between variety and inoculation status (p-value$=0.440$ from F-stat$=0.999$ on $8$ and $135$ df).
 \end{quote}
 <<refitno2way, echo=FALSE, results='asis'>>=
-glm2way2 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt
+glm2way2 <- glm(cbind(infected, uninfected) ~ block*variety + variety*n.trt
                + variety*status2,
@@
 <<refitno2way3, echo=FALSE, results='asis', include=FALSE>>=
-glm2way3 <- glm(cbind(infected, total) ~ block*variety + variety*n.trt,
+glm2way3 <- glm(cbind(infected, uninfected) ~ block*variety + variety*n.trt,
                family = quasibinomial, data = fert.sim1)
@@
 \begin{quote}
-There is no evidence that the odds of virus infection depends on nitrogen application time after controlling for block, variety, inoculation status, and the $block*variety$ interaction (p-value$=0.612$ from F-stat$=0.49$ on $2$ and $147$ df).
+There is no evidence that the odds of virus infection depends on nitrogen application time after controlling for block, variety, inoculation status, and the $block*variety$ interaction (p-value$=0.633$ from F-stat$=0.459$ on $2$ and $147$ df).
 \end{quote}
@@
 <<glmnointeractions, echo=TRUE, results='hide'>>=
-glmnointeractions <- glm(cbind(infected, total) ~ block*variety + variety + n.trt + status2,
+glmnointeractions <- glm(cbind(infected, uninfected) ~ block*variety + variety + n.trt + status2,
                family = quasibinomial, data = fert.sim1)
@@
 <<dropindevftest, echo=FALSE>>=
-glm.no3way <- glm(cbind(infected, total) ~ block*variety + variety*n.trt + n.trt*status2 + variety*status2,
+glm.no3way <- glm(cbind(infected, uninfected) ~ block*variety + variety*n.trt + n.trt*status2 + variety*status2,
                family = quasibinomial, data = fert.sim1)
@@
 \begin{quote}
-For inoculated plants of variety $1$, the odds of virus infection when nitrogen is applied in the early spring is estimated to be $3.01$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from  $0.295$ to $104.9$ times.
-For inoculated plants of variety $1$, the odds of virus infection when nitrogen is applied in the late spring is estimated to be $5.05$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from $0.652$ to $166.4$ times.
+For inoculated plants of variety $1$, the odds of virus infection when nitrogen is applied in the early spring is estimated to be $3.04$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from  $0.264$ to $136.6$ times.
+For inoculated plants of variety $1$, the odds of virus infection when nitrogen is applied in the late spring is estimated to be $5.13$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from $0.604$ to $218.4$ times.
 \end{quote}
@@
 #then fit model with var3
-glm.refvar3 <- glm(cbind(infected, total) ~ block/var3 + var3*n.trt*status2,
+glm.refvar3 <- glm(cbind(infected, uninfected) ~ block/var3 + var3*n.trt*status2,
                family = quasibinomial, data = fert.sim1)
@@
 #then fit model with var3
-glm.refvar3cntl <- glm(cbind(infected, total) ~ block/var3 + var3*n.trt*statuscntl,
+glm.refvar3cntl <- glm(cbind(infected, uninfected) ~ block/var3 + var3*n.trt*statuscntl,
                family = quasibinomial, data = fert.sim1)
@@
 \begin{quote}
-For inoculated plants of variety $3$, the odds of virus infection when nitrogen is applied in the early spring is estimated to be about equal to the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from  $0.191$ to $5.28$ times.
-For inoculated plants of variety $3$, the odds of virus infection when nitrogen is applied in the late spring is estimated to be $1.24$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from $0.269$ to $6.201$ times.
+For inoculated plants of variety $3$, the odds of virus infection when nitrogen is applied in the early spring is estimated to be about equal to the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from  $0.174$ to $5.763$ times.
+For inoculated plants of variety $3$, the odds of virus infection when nitrogen is applied in the late spring is estimated to be $1.26$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from $0.252$ to $6.910$ times.
 \end{quote}
@@
 \begin{quote}
-For non-inoculated plants of variety $3$, the odds of virus infection when nitrogen is applied in the early spring is estimated to be $2.3$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from  $0.797$ to $7.72$ times.
-For non-inoculated plants of variety $3$, the odds of virus infection when nitrogen is applied in the late spring is estimated to be $1.45$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from $0.446$ to $5.16$ times.
+For non-inoculated plants of variety $3$, the odds of virus infection when nitrogen is applied in the early spring is estimated to be $2.49$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from  $0.817$ to $8.965$ times.
+For non-inoculated plants of variety $3$, the odds of virus infection when nitrogen is applied in the late spring is estimated to be $1.54$ times the odds of virus infection when nitrogen is applied in the fall, with a $95\%$ likelihood-based confidence interval from $0.444$ to $5.87$ times.
 \end{quote}
